---
title: "psychencode_DE_analyses"
author: "Daniel Kiss"
date: "7/14/2023"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(Seurat)
  library(edgeR)
  library(ggrepel)
  library(cowplot)
  library(limma)
  library(ggpubr)
  library(gprofiler2)
  library(EnhancedVolcano)
  library(patchwork)
})
theme_set(theme_classic2())
#Colour palette
cbPalette <- c("#56B4E9", "#009E73","#E69F00", "#0072B2", "#D55E00", "#CC79A7","#000000","#F0E442")
```

Load Data
```{r}
pseudobulk = read.delim("/external/rprshnas01/netdata_kcni/stlab/SZBDMulticohort/pseudobulk_fine_grained.counts.tsv")
pseudobulk_0 = pseudobulk
pseudobulk_metadata = read.delim("/external/rprshnas01/netdata_kcni/stlab/SZBDMulticohort/pseudobulk_fine_grained.metadata.tsv")
```

Subset and Format Data
```{r}
#Add group collumn for comnparisons
pseudobulk_metadata = pseudobulk_metadata %>%
  mutate(group = case_when(
    Phenotype == "SZ" & Age >= 70 ~ "SCZ_Over70",
    Phenotype == "SZ" & Age < 70 ~ "SCZ_Under70",
    Phenotype == "CON" & Age >= 70 ~ "CON_Over70",
    Phenotype == "CON" & Age < 70 ~ "CON_Under70",
    TRUE ~ NA_character_
  )) %>%
  #Add new gender and ID columns
  mutate(newID = paste(unique_donor_ID, group, sep = "_")) %>%
  mutate(newGender = paste(Gender, group, sep = "_"))

#Convert psuedobulk to cpm
pseudobulk_cpm = pseudobulk[,3:ncol(pseudobulk)] %>% lapply(as.numeric) %>% as.data.frame() %>% cpm() %>% as.data.frame()
pseudobulk_first_cols = pseudobulk[,1:2]
pseudobulk_cpm =pseudobulk_first_cols %>% cbind(pseudobulk_cpm)
#Combine cpm matrix with metadata
pseudobulk_combined = left_join(pseudobulk_metadata, pseudobulk_cpm)

```

Define subset variables and model variables
```{r}
USE_CELL_TYPE = "In-SST"
USE_COHORT = 'McLean'

# Identify subjects with information for disease, sex, PMI, and age
use_subjects = pseudobulk_metadata %>% 
  filter(Celltype == USE_CELL_TYPE, Cohort == USE_COHORT, num_cells > 50) %>% 
  pull(unique_donor_ID)

disease = factor( pseudobulk_metadata %>% 
                   filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
                   pull(Phenotype), levels = c('CON', 'SZ'))
sex =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
  pull(Gender)
pmi =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
  pull(PMI)
age =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
  pull(Age)
log_10_cells_per_donor =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
  pull(num_cells) %>% log10()
#Grouping variable added earlier 
group = factor( pseudobulk_metadata %>% 
                   filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% pull(group))
```

Extract the pseudobulk counts for chosen cells of the selected subjects
```{r}
pseudobulk_single_celltype_counts = pseudobulk %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE)

pseudobulk_single_celltype_counts =pseudobulk_single_celltype_counts[, 3:ncol(pseudobulk_single_celltype_counts)] %>% t() %>% as.data.frame()
colnames(pseudobulk_single_celltype_counts) = pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
  pull(unique_donor_ID)
```

DGEList Object Setup
```{r}
dge0 = DGEList(pseudobulk_single_celltype_counts, genes = row.names(pseudobulk_single_celltype_counts))

min_samples_expressing_gene <- pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
  nrow * 0.5 # The 0.8 here refers to the fraction of total samples that needs to express the gene
dge0 = dge0[rowSums(dge0$counts >= 1) >= min_samples_expressing_gene, ] # Filter genes that are detected in at least 80% of samples

dge0 = calcNormFactors(dge0, method = "TMM")

### Set up design based on the factors defined above
design = model.matrix(~0 + group + age + pmi + sex + log_10_cells_per_donor) 

# Perform voom transformation
vm = voom(dge0, design, plot = TRUE)

# Perform lmFit and eBayes steps
fit = lmFit(vm, design)
fit = eBayes(fit)

```

Plot genes associated with sex - to make sure analysis was done correctly thus far
```{r}
deg_table_sex = topTable(fit, coef = "sexMale",  n = Inf, sort = "none", adjust.method = "BH") 

deg_table_sex %>% 
  ggplot(aes(x = logFC, y = -log10(P.Value), label = genes)) + 
  geom_point() +
  geom_text_repel(data = subset(deg_table_sex, adj.P.Val < 0.05), 
                  vjust = 1.5) +
  geom_point(data = subset(deg_table_sex, adj.P.Val < 0.05), 
             aes(color = 'red')) + 
  theme(legend.position = "none")

```

Now let's move on to group-specific comparisons

## Under 70: CON vs. SCZ
```{r}
#Create Contrasts
contr = makeContrasts(groupCON_Under70 - groupSCZ_Under70 , levels = colnames(coef(fit)))
#Estimate contrast for each gene
tmp <- contrasts.fit(fit, contr)
#Empirical Bayes smoothing of standard errors (shrinks standard errors that are much larger or smaller than those from other genes towards the average standard error) (see https://www.degruyter.com/doi/10.2202/1544-6115.1027)
tmp <- eBayes(tmp)
#What genes are most differentially expressed?
DE_under_70 <- topTable(tmp, sort.by = "P", n = Inf)
#How many significant DE genes are there?
length(which(DE_under_70$adj.P.Val < 0.05))

#Plot
EnhancedVolcano(DE_under_70,
                lab = rownames(DE_under_70),
                x = 'logFC',
                y = 'P.Value')
```


## Over 70: CON vs. SCZ
```{r}
#Create Contrasts
contr = makeContrasts(groupCON_Over70 - groupSCZ_Over70 , levels = colnames(coef(fit)))
#Estimate contrast for each gene
tmp <- contrasts.fit(fit, contr)
#Empirical Bayes smoothing of standard errors (shrinks standard errors that are much larger or smaller than those from other genes towards the average standard error) (see https://www.degruyter.com/doi/10.2202/1544-6115.1027)
tmp <- eBayes(tmp)
#What genes are most differentially expressed?
DE_over_70 <- topTable(tmp, sort.by = "P", n = Inf)
#How many significant DE genes are there?
length(which(DE_over_70$adj.P.Val < 0.05))

#Plot
EnhancedVolcano(DE_over_70,
                lab = rownames(DE_over_70),
                x = 'logFC',
                y = 'P.Value')
```

Plot t-statistics from both DE comparisons
```{r}
tplot = merge(DE_over_70, DE_under_70, by = 0) %>% 
  ggplot(aes(x = t.x, y = t.y)) + 
  geom_smooth(method = loess) +
  geom_point(size = 1) +
  ylab("Over 70") +
  xlab("Under 70") +
  ggtitle("t-Statistics")

tplot 

cor.test(DE_over_70$t, DE_under_70$t)
```

Repeat analysis without contrasts - to see if it makes a difference:
```{r}
#Bin subjectID by age
Under70 = pseudobulk_metadata %>% filter(Celltype == USE_CELL_TYPE, Cohort == USE_COHORT, Age < 70, num_cells > 50) %>%
  pull(unique_donor_ID)
Over70 = pseudobulk_metadata %>% filter(Celltype == USE_CELL_TYPE, Cohort == USE_COHORT, Age >= 70, num_cells > 50) %>%
  pull(unique_donor_ID)
```

Extract the pseudobulk counts for chosen cells of the selected subjects
```{r}
#Under70
pseudobulk_single_celltype_counts_Under70 = pseudobulk %>% 
  filter(unique_donor_ID %in% Under70, Celltype == USE_CELL_TYPE)

pseudobulk_single_celltype_counts_Under70 =pseudobulk_single_celltype_counts_Under70[, 3:ncol(pseudobulk_single_celltype_counts_Under70)] %>% t() %>% as.data.frame()
colnames(pseudobulk_single_celltype_counts_Under70) = pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% Under70, Celltype == USE_CELL_TYPE) %>% 
  pull(unique_donor_ID)

#Over70
pseudobulk_single_celltype_counts_Over70 = pseudobulk %>% 
  filter(unique_donor_ID %in% Over70, Celltype == USE_CELL_TYPE)

pseudobulk_single_celltype_counts_Over70 =pseudobulk_single_celltype_counts_Over70[, 3:ncol(pseudobulk_single_celltype_counts_Over70)] %>% t() %>% as.data.frame()
colnames(pseudobulk_single_celltype_counts_Over70) = pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% Over70, Celltype == USE_CELL_TYPE) %>% 
  pull(unique_donor_ID)
```

Define subset variables and model variables - Under 70
```{r}
# Identify subjects with information for disease, sex, PMI, and age
use_subjects = pseudobulk_metadata %>% 
  filter(Celltype == USE_CELL_TYPE, Cohort == USE_COHORT, num_cells > 50, Age < 70) %>% 
  pull(unique_donor_ID)

disease = factor( pseudobulk_metadata %>% 
                   filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
                   pull(Phenotype), levels = c('CON', 'SZ'))
sex =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
  pull(Gender)
pmi =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
  pull(PMI)
age =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
  pull(Age)
log_10_cells_per_donor =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
  pull(num_cells) %>% log10()
```


DGEList Object Setup - Under 70
```{r}
dge0 = DGEList(pseudobulk_single_celltype_counts_Under70, genes = row.names(pseudobulk_single_celltype_counts_Under70))

min_samples_expressing_gene <- pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
  nrow * 0.5 # The 0.8 here refers to the fraction of total samples that needs to express the gene
dge0 = dge0[rowSums(dge0$counts >= 1) >= min_samples_expressing_gene, ] # Filter genes that are detected in at least 80% of samples

dge0 = calcNormFactors(dge0, method = "TMM")

### Set up design based on the factors defined above
design = model.matrix(~disease + age + pmi + sex + log_10_cells_per_donor) 

# Perform voom transformation
vm = voom(dge0, design, plot = TRUE)

# Perform lmFit and eBayes steps
fit = lmFit(vm, design)
fit = eBayes(fit)

DE_under_70 <- topTable(fit, coef = "diseaseSZ",  n = Inf, sort = "none", adjust.method = "BH") 
#How many significant DE genes are there?
length(which(DE_under_70$adj.P.Val < 0.05))

#Plot
EnhancedVolcano(DE_under_70,
                lab = rownames(DE_under_70),
                x = 'logFC',
                y = 'P.Value')
```


Define subset variables and model variables - Over 70
```{r}
# Identify subjects with information for disease, sex, PMI, and age
use_subjects = pseudobulk_metadata %>% 
  filter(Celltype == USE_CELL_TYPE, Cohort == USE_COHORT, num_cells > 50, Age >= 70) %>% 
  pull(unique_donor_ID)

disease = factor( pseudobulk_metadata %>% 
                   filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
                   pull(Phenotype), levels = c('CON', 'SZ'))
sex =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
  pull(Gender)
pmi =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
  pull(PMI)
age =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
  pull(Age)
log_10_cells_per_donor =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
  pull(num_cells) %>% log10()
```


DGEList Object Setup - Over 70
```{r}
dge0 = DGEList(pseudobulk_single_celltype_counts_Over70, genes = row.names(pseudobulk_single_celltype_counts_Over70))

min_samples_expressing_gene <- pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% use_subjects, Celltype == USE_CELL_TYPE) %>% 
  nrow * 0.5 # The 0.8 here refers to the fraction of total samples that needs to express the gene
dge0 = dge0[rowSums(dge0$counts >= 1) >= min_samples_expressing_gene, ] # Filter genes that are detected in at least 80% of samples

dge0 = calcNormFactors(dge0, method = "TMM")

### Set up design based on the factors defined above
design = model.matrix(~disease + age + pmi + sex + log_10_cells_per_donor) 

# Perform voom transformation
vm = voom(dge0, design, plot = TRUE)

# Perform lmFit and eBayes steps
fit = lmFit(vm, design)
fit = eBayes(fit)

DE_over_70 <- topTable(fit, coef = "diseaseSZ",  n = Inf, sort = "none", adjust.method = "BH") 
#How many significant DE genes are there?
length(which(DE_over_70$adj.P.Val < 0.05))

#Plot
EnhancedVolcano(DE_over_70,
                lab = rownames(DE_over_70),
                x = 'logFC',
                y = 'P.Value')
```

```{r}
tplot = merge(DE_over_70, DE_under_70, by = 0) %>% 
  ggplot(aes(x = t.x, y = t.y)) + 
  geom_smooth(method = loess) +
  geom_point(size = 1) +
  ylab("Over 70") +
  xlab("Under 70") +
  ggtitle("t-Statistics")

tplot 

cor.test(DE_over_70$t, DE_under_70$t)
```

-----------------------------------------------------

Proceeding with no-contrasts for subsequent analyses

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(Seurat)
  library(edgeR)
  library(ggrepel)
  library(cowplot)
  library(limma)
  library(ggpubr)
  library(gprofiler2)
  library(EnhancedVolcano)
})
theme_set(theme_classic2())
#Colour palette
cbPalette <- c("#56B4E9", "#009E73","#E69F00", "#0072B2", "#D55E00", "#CC79A7","#000000","#F0E442")
```
>>>>>>> 5ba7279da9217526bdb401774fe434181dc5c35b

Load Data // Define Variables
```{r}
pseudobulk = read.delim("/external/rprshnas01/netdata_kcni/stlab/SZBDMulticohort/pseudobulk_fine_grained.counts.tsv")
pseudobulk_metadata = read.delim("/external/rprshnas01/netdata_kcni/stlab/SZBDMulticohort/pseudobulk_fine_grained.metadata.tsv")
CELL_TYPES = pseudobulk_metadata$Celltype %>% unique()
USE_COHORT = 'MtSinai'
USE_COHORT = 'McLean'
```


Create a list of relevant data for each cell type
```{r}
DE_data = list()
#Loop through ALL cell types and perform above analyses
for(CELL_TYPE in CELL_TYPES) {
  #Define subjects based on cell type, cohort, and min cell number
  under_70_subjects = pseudobulk_metadata %>% filter(Celltype == CELL_TYPE, Cohort == USE_COHORT, num_cells > 50, Age <= 70) %>% 
  pull(unique_donor_ID)
  
  over_70_subjects = pseudobulk_metadata %>% 
  filter(Celltype == CELL_TYPE, Cohort == USE_COHORT, num_cells > 50, Age > 70) %>% 
  pull(unique_donor_ID)

  #Skip cell type if no subjects make it past filtering
  if(length(under_70_subjects) <= 10 || length(over_70_subjects) <= 10) {
    next  # Skip to the next iteration of the loop
  }
  #Define model variables for under 70
  disease = factor( pseudobulk_metadata %>% filter(unique_donor_ID %in% under_70_subjects, Celltype == CELL_TYPE) %>% 
    pull(Phenotype), levels = c('CON', 'SZ'))
  sex =  pseudobulk_metadata %>% filter(unique_donor_ID %in% under_70_subjects, Celltype == CELL_TYPE) %>% 
    pull(Gender)
  pmi =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% under_70_subjects, Celltype == CELL_TYPE) %>% 
  pull(PMI)
  age =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% under_70_subjects, Celltype == CELL_TYPE) %>% 
  pull(Age)
  log_10_cells_per_donor =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% under_70_subjects, Celltype == CELL_TYPE) %>% 
  pull(num_cells) %>% log10()
  #Create model matrix 
  design_under_70 = model.matrix(~disease + age + pmi + sex + log_10_cells_per_donor) 
  
    #Define model variables for over 70
  disease = factor( pseudobulk_metadata %>% filter(unique_donor_ID %in% over_70_subjects, Celltype == CELL_TYPE) %>% 
    pull(Phenotype), levels = c('CON', 'SZ'))
  sex =  pseudobulk_metadata %>% filter(unique_donor_ID %in% over_70_subjects, Celltype == CELL_TYPE) %>% 
    pull(Gender)
  pmi =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% over_70_subjects, Celltype == CELL_TYPE) %>% 
  pull(PMI)
  age =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% over_70_subjects, Celltype == CELL_TYPE) %>% 
  pull(Age)
  log_10_cells_per_donor =  pseudobulk_metadata %>% 
  filter(unique_donor_ID %in% over_70_subjects, Celltype == CELL_TYPE) %>% 
  pull(num_cells) %>% log10()
  #Create model matrix 
  design_over_70 = model.matrix(~disease + age + pmi + sex + log_10_cells_per_donor) 
  
  #Create count matrices
  under_70 = pseudobulk_metadata %>% filter(Age <= 70, unique_donor_ID %in% under_70_subjects, Celltype == CELL_TYPE) %>% pull(unique_donor_ID) 
  counts_under_70 = pseudobulk %>% filter(unique_donor_ID %in% under_70, Celltype == CELL_TYPE)
  counts_under_70 = counts_under_70[, 3:ncol(counts_under_70)] %>% t() %>% as.data.frame()
  colnames(counts_under_70) = under_70
  over_70 = pseudobulk_metadata %>% filter(Age > 70, unique_donor_ID %in% over_70_subjects, Celltype == CELL_TYPE) %>% pull(unique_donor_ID) 
  counts_over_70 = pseudobulk %>% filter(unique_donor_ID %in% over_70, Celltype == CELL_TYPE)
  counts_over_70 = counts_over_70[, 3:ncol(counts_over_70)] %>% t() %>% as.data.frame()
  colnames(counts_under_70) = under_70
  #Create DGE objects
  dge_under = DGEList(counts_under_70, genes = row.names(counts_under_70))
  min_samples <- pseudobulk_metadata %>% filter(unique_donor_ID %in% under_70_subjects, Celltype == CELL_TYPE) %>% nrow * 0.8
  dge_under = dge_under[rowSums(dge_under$counts >= 1) >= min_samples, ]
  dge_under = calcNormFactors(dge_under, method = "TMM")
  dge_over = DGEList(counts_over_70, genes = row.names(counts_over_70))
  min_samples <- pseudobulk_metadata %>% filter(unique_donor_ID %in% over_70_subjects, Celltype == CELL_TYPE) %>% nrow * 0.8
  dge_over = dge_over[rowSums(dge_over$counts >= 1) >= min_samples, ]
  dge_over = calcNormFactors(dge_over, method = "TMM")
  
  #Create lists of data for each group
  under_70_data = list(design_under_70, dge_under)
  names(under_70_data) = c("design", "dge")
  over_70_data = list(design_over_70, dge_over)
  names(over_70_data) = c("design", "dge")
  joined_data = list(under_70_data, over_70_data)
  names(joined_data) = c("under_70", "over_70")
  
  DE_data[[CELL_TYPE]] = joined_data
}
#We now have a nested list with ALL data required for DE analysis for all cell types with >1 subject per age group
```

Simple volcano plot function 
```{r}
simple_volcano <- function(x) {
  x %>%
    ggplot(aes(x = logFC, y = -log10(P.Value), label = genes)) + 
    geom_point() +
    geom_text_repel(data = subset(x, adj.P.Val < 0.1), 
                    aes(label = genes), 
                    vjust = 1.5) + 
    geom_point(data = subset(x, adj.P.Val < 0.1), 
               aes(color = 'red')) + 
    theme_cowplot() +
    theme(legend.position = "none")
}
```


Now that we have all data pre-processed and organized for DE analysis, let's loop through DE_data and output results to DE_results (This will take a few minutes to run)
```{r}
DE_results = list()
for(CELL_TYPE in names(DE_data)) {
  #Load data from DE_data
  dge_under = DE_data[[CELL_TYPE]]$under_70$dge 
  dge_over = DE_data[[CELL_TYPE]]$over_70$dge 
  design_under = DE_data[[CELL_TYPE]]$under_70$design
  design_over = DE_data[[CELL_TYPE]]$over_70$design
  #Voom transform
  vm_under = voom(dge_under, design_under, plot = F) 
  fit_under = lmFit(vm_under, design_under) %>% eBayes()
  DE_under = topTable(fit_under, coef = "diseaseSZ",  n = Inf, sort = "none", adjust.method = "BH") 
  vm_over = voom(dge_over, design_over, plot = FALSE) 
  fit_over = lmFit(vm_over, design_over) %>% eBayes()
  DE_over = topTable(fit_over, coef = "diseaseSZ", n = Inf, sort = "none", adjust.method = "BH")
  #Plots etc.
  volcano_under = DE_under %>% simple_volcano()
  volcano_over = DE_over %>% simple_volcano()
  tplot = merge(DE_under, DE_over, by = 0) %>% 
    ggplot(aes(x = t.x, y = t.y)) + 
    geom_smooth(method = loess) +
    geom_point(size = 1) +
    ylab("Over 70") +
    xlab("Under 70") +
    ggtitle(paste0(CELL_TYPE, " t-Statistics"))
  
  #Append to list
  joined_results = list(DE_under, DE_over)
  names(joined_results) = c("DE_under", "DE_over")
  joined_volcano = list(volcano_under, volcano_over)
  names(joined_volcano) = c("Under 70", "Over 70")
  joined_all = list(joined_results, joined_volcano, tplot)
  names(joined_all) = c("DE_tables", "Volcano_Plots", "t_Plot")
  DE_results[[CELL_TYPE]] = joined_all
}
```

We will run the code above 2x, once for McLean and once for Sinai, then save each result for later use 
```{r}
saveRDS(DE_data, file = paste0("/external/rprshnas01/kcni/dkiss/cell_prop_psychiatry/data/DE_data_", USE_COHORT))
saveRDS(DE_results, file = paste0("/external/rprshnas01/kcni/dkiss/cell_prop_psychiatry/data/DE_results_", USE_COHORT))
```



Create a data frame of pearson correlation data between under 70 and over 70 groups for each cell type
```{r}
#Load data and initialize empty data frame
DE_results_MtSinai = readRDS("/external/rprshnas01/kcni/dkiss/cell_prop_psychiatry/data/DE_results_MtSinai")
DE_results_McLean = readRDS("/external/rprshnas01/kcni/dkiss/cell_prop_psychiatry/data/DE_results_McLean")
#Compute pearson correlation values and save them to vectors with NA if the cell type was filtered out
mclean_corr = c()
mtsinai_corr = c()
for(CELL_TYPE in CELL_TYPES) {
  if(DE_results_McLean[[CELL_TYPE]] %>% is.null == F) {
    tplot = DE_results_McLean[[CELL_TYPE]][['t-Plot']]
    correlation <- cor(tplot$data$t.x, tplot$data$t.y)
    mclean_corr = mclean_corr %>% append(correlation)
  }
  else {
    mclean_corr = mclean_corr %>% append(NA)
  }
}
for(CELL_TYPE in CELL_TYPES) {
  if(DE_results_MtSinai[[CELL_TYPE]] %>% is.null == F) {
    tplot = DE_results_McLean[[CELL_TYPE]][['t-Plot']]
    correlation <- cor(tplot$data$t.x, tplot$data$t.y)
    mtsinai_corr = mtsinai_corr %>% append(correlation)
  }
  else {
    mtsinai_corr = mtsinai_corr %>% append(NA)
  }
}
#Create data frame
corr_data = data.frame(cell_type = CELL_TYPES, mtsinai_corr = mtsinai_corr, mclean_corr = mclean_corr)  %>%
  pivot_longer(cols = mtsinai_corr:mclean_corr, names_to = 'study', values_to = 'corr') %>%  
  mutate(class = case_when(
    str_detect(cell_type, "In") ~ "Inhibitory",
    str_detect(cell_type, "Ex") ~ "Excitatory",
    T ~ "Non-Neuronal"))
  

```

Let's plot corr_data
```{r, fig.height=4, fig.width=8 }
corr_plot = corr_data %>%
  ggplot(aes(x = cell_type, y = corr, fill = study)) + 
  geom_bar(stat = "identity", position = "dodge") +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 13), 
    axis.title.x = element_text(vjust = 6, size = 16),
    axis.title.y = element_text(size = 16),
    strip.text.x = element_text(size = 16)) + 
  facet_grid(~class, drop = T, scale = "free_x", space = "free") + 
  ylab('Pearson Correlation - Above/Below 70') + 
  xlab('Cell Type') 

corr_plot
```

Next: Plot t-plots for each cohort -> will take a while to run
```{r, fig.height= 10, fig.width= 15}
# Create an empty list to store the tplot ggplot objects
combined_tplots <- list()
for (CELL_TYPE in names(DE_results)) {
  tplot <- DE_results[[CELL_TYPE]]$t_Plot
  combined_tplots[[CELL_TYPE]] <- tplot
}
# Combine tplot ggplot objects
combined_tplots <- cowplot::plot_grid(plotlist = combined_tplots, ncol = 5)

combined_tplots


```



