---
title: "psychencode_supplemental_figures"
author: "Daniel Kiss"
date: "3/7/2024"
output: html_document
---

Packages etc.
```{r}
library(dplyr)
library(tidyverse)
library(edgeR)
library(markerGeneProfile) 
library(matrixStats)
library(cowplot)
library(broom)
library(knitr)
library(ggpubr)
library(ggrepel)
library(patchwork)
library(ggsignif)
library(modelr)
library(ggbeeswarm)
library(lemon)
theme_set(theme_classic2())
#Colour palette
cbPalette = c("#E69F00", "#D55E00", "#CC79A7")
```

```{r}
#Add PMI info from CMC metadata
psychencode_metadata = read.csv(("/external/rprshnas01/external_data/psychencode/PsychENCODE/Metadata/CapstoneCollection_Metadata_Clinical.csv"))
CMC_metadata = read.csv("/external/rprshnas01/external_data/psychencode/PsychENCODE/CMC/Metadata/SYNAPSE_TABLE_QUERY_123020650.csv") %>% filter(dataType == "geneExpression", fileFormat == "tsv")  %>% left_join(psychencode_metadata)
names(CMC_metadata)[names(CMC_metadata) == "Individual_ID"] = "individualID"

#Cell type proportions for each data set merged with metadata - ready for modelling 
acc_estimations = read_csv("/external/rprshnas01/kcni/dkiss/cell_prop_psychiatry/data/cmc_acc_mgp_estimations.csv") %>% filter(primaryDiagnosis %in% c("control", "Schizophrenia")) %>% left_join(CMC_metadata[, c("individualID", "PMI")], by = "individualID")
acc_estimations$newStudy = acc_estimations$newStudy %>% factor(levels = c("Pitt_ACC", "MSSM_ACC", "Penn_ACC"))
acc_estimations_long = acc_estimations %>% pivot_longer(cols = Astrocyte:VLMC, names_to = 'cell_type', values_to = 'rel_prop') %>%
  mutate(cell_class = case_when(
    cell_type %in% c("PVALB", "SST", "VIP", "LAMP5", "PAX6") ~ "Inhibitory",
    cell_type %in% c("IT", "L4.IT", "L5.ET", "L5.6.IT.Car3", "L5.6.NP", "L6.CT", "L6b") ~ "Excitatory",
    TRUE ~ "Non-Neuronal"
  ))
acc_estimations_long$cell_class = acc_estimations_long$cell_class %>% factor(levels = c("Inhibitory", "Excitatory", "Non-Neuronal"))

```

Cohort demographic summary
```{r}
acc_summary_df = acc_estimations %>%
    distinct(individualID, .keep_all = TRUE) %>% # Ensure rows are unique based on individualID
    group_by(newStudy) %>%
    summarise(
        CohortName = first(newStudy),
        PercentMale = mean(reportedGender == "male", na.rm = TRUE) * 100,
        AvgAge = mean(ageDeath, na.rm = TRUE),
        AgeSD = sd(ageDeath, na.rm = TRUE),
        NumCases = sum(primaryDiagnosis == "Schizophrenia", na.rm = TRUE),
        NumControls = sum(primaryDiagnosis == "control", na.rm = TRUE)
    )
```

Loading LCM proportions
```{r}
#Load Sibille Lab microscopy
sibille_lab_cell_props = sibille_lab_cell_fractions = read_csv('/external/rprshnas01/netdata_kcni/stlab/cross_cohort_MGPs/sibille_lab_cell_fractions.csv') %>% filter(Subject.Group %in% c("Control", "SCHIZ")) %>% distinct()

# this is the tetrad metadata
tetrad_path = '/external/rprshnas01/netdata_kcni/stlab/Collaborators/Sibille/RNAseq_processed/'
tetrad_meta = read_csv(paste0(tetrad_path, "pTable with correct med info.csv"))
tetrad_meta$sample_names = str_sub(tetrad_meta$X, end=-5) %>% make.names()
tetrad_meta$sample_names =  gsub("_", ".", tetrad_meta$sample_names)
tetrad_meta = tetrad_meta %>% mutate(Subject = HU., 
                       Reported.Gender = case_when(Sex == "M" ~"Male",
                                                   Sex == "F" ~ "Female"),
                       Age_norm = Age, 
                       PMI..in.hours. = PMI) %>% filter(Cell.Type == "SST")

microCTPs = left_join(sibille_lab_cell_props, tetrad_meta) %>% filter(Subject != 1367)
microCTPs_long =  microCTPs %>% pivot_longer(cols = c(PVALB, PYR_23, PYR_56, SST, VIP), names_to = 'cell_type', values_to = 'rel_prop') 
microCTPs_long$rel_prop = microCTPs_long$rel_prop %>% replace(is.na(.), 0)

```

Load LCM DE table
```{r}
scz_lcm_seq_de = readxl::read_excel("~/cell_prop_psychiatry/data/psychencode_lcm_DE_scz.xlsx")
#Calculate 95% CI
scz_lcm_seq_de <- scz_lcm_seq_de %>%
  mutate(
    SE = logFC / t,  
    CI.L = logFC - 1.96 * SE, 
    CI.R = logFC + 1.96 * SE   
  )
```


Model rCTPs based on diagnosis and covariates separately in each cohort
```{r}
acc_combined_lms = acc_estimations_long %>%
  group_by(newStudy, cell_type) %>%
  do(tidy(lm(scale(rel_prop) ~ scale(RIN) + scale(ageDeath) + scale(PMI) +
               reportedGender + primaryDiagnosis, data = .))) %>%
  ungroup() %>%
  mutate(padj = p.adjust(p.value, method = 'fdr'),
         term = recode(term,
                       `(Intercept)` = "Intercept",
                       `reportedGendermale` = "gender:Male",
                       `primaryDiagnosisSchizophrenia` = "SCZ",
                       `scale(ageDeath)` = "Age",
                       `scale(PMI)` = "PMI",
                       `scale(RIN)` = "RIN")) %>%
  merge(unique(acc_estimations_long[c('dataset', 'newStudy')]), by.x = "newStudy") %>%
  merge(unique(acc_estimations_long[c('cell_class', 'cell_type')]), by.x = "cell_type") 

acc_combined_lms$newStudy = acc_combined_lms$newStudy %>% factor(levels = c("Pitt_ACC", "MSSM_ACC", "Penn_ACC"))
```

Interaction mega-analysis w/ all data 
```{r}
acc_mega_lms = acc_estimations_long %>% 
  group_by(cell_type) %>%
  do(tidy(lm(scale(rel_prop) ~ scale(RIN) + scale(ageDeath) + scale(PMI) +
             reportedGender + primaryDiagnosis + newStudy + scale(ageDeath)*primaryDiagnosis, data = .))) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(`p.value`, method = 'fdr'),
         term = recode(term,
                       `(Intercept)` = "Intercept",
                       `reportedGendermale` = "gender:Male",
                       `primaryDiagnosisSchizophrenia` = "SCZ",

                       `scale(ageDeath)` = "Age",
                       `scale(ageDeath):primaryDiagnosisSchizophrenia` = "Age x Diagnosis",
                       `scale(PMI)` = "PMI", 
                       `newStudyNIMH_HBCC` = "NIMH_HBCC",
                       `newStudyLIBD_szControl` = "LIBD_szControl",
                       `newStudyPitt` = "Pitt_ACC",
                       `newStudyMSSM` = "MSSM_ACC",
                       `newStudyPenn` = "Penn_ACC",
                       `scale(RIN)` = "RIN")) %>%
  merge(unique(acc_estimations_long[c('cell_class', 'cell_type')]), by.x = "cell_type") 
```

Functions to  generate age range bins 
```{r}
# Create function to map age to age range
age_to_range = function(age) {
    if(age >= 15 & age < 20) {
    return("15-19")
  } else {
    lower_bound = floor((age - 20) / 10) * 10 + 20
    upper_bound = lower_bound + 9
    return(paste0(lower_bound, "-", upper_bound))
  }
}
# Create function to bin ages to >=70 and <70
classify_ages = function(ages) {
  classifications = ifelse(ages <= 70, "Under 70", "Over 70")
  return(classifications)
}

# Create function to map age to age range based on 20% quantiles
age_to_range_quantile <- function(age) {
  quantiles = quantile(age, probs = seq(0.5, 1, 0.5))
  range_vec = c()
  for (i in 1:length(quantiles)) {
    if (i == 1) {
      range_vec[age <= quantiles[i]] <- paste0(0, "-", quantiles[i])
    } else {
      range_vec[age > quantiles[i-1] & age <= quantiles[i]] <- paste0(quantiles[i-1], "-", quantiles[i])
    }
  }
  range_vec
}
```

Create age groups from age 20-90 in increments of 10 years. Plot meganalysis beta coefs against age group.
```{r}
acc_estimations_long$age_range = sapply(acc_estimations_long$ageDeath, age_to_range)

acc_lms_by_age = acc_estimations_long %>% filter(ageDeath >= 30) %>%
  group_by(age_range, cell_type) %>%
  do(tidy(lm(scale(rel_prop) ~ scale(RIN) + scale(PMI) +
             reportedGender + primaryDiagnosis + newStudy, data = .))) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(`p.value`, method = 'fdr'),
         term = recode(term,
                       `(Intercept)` = "Intercept",
                       `reportedGendermale` = "gender:Male",
                       `primaryDiagnosisSchizophrenia` = "SCZ",
                       `scale(ageDeath)` = "Age",
                       `scale(PMI)` = "PMI", 
                       `newStudyNIMH_HBCC` = "NIMH_HBCC",
                       `newStudyPitt` = "Pitt_ACC",
                       `newStudyMSSM` = "MSSM_ACC",
                       `newStudyPenn` = "Penn_ACC",
                       `scale(RIN)` = "RIN")) %>%
  merge(unique(acc_estimations_long[c('cell_class', 'cell_type')]), by.x = "cell_type") 
```



Bulk RNAseq Meganalysis based on age +/- 70
```{r}
acc_estimations_long$age_class <- sapply(acc_estimations_long$ageDeath, classify_ages) %>% factor(levels = c("Under 70", "Over 70"))
acc_mega_lms_by_age_class = acc_estimations_long %>% 
  group_by(age_class, cell_type) %>%
  do(tidy(lm(scale(rel_prop) ~ scale(RIN) + scale(PMI) +
             reportedGender + newStudy + primaryDiagnosis, data = .))) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(`p.value`, method = 'fdr'),
         term = recode(term,
                       `(Intercept)` = "Intercept",
                       `reportedGendermale` = "gender:Male",
                       `primaryDiagnosisSchizophrenia` = "SCZ",
                       `scale(ageDeath)` = "Age",
                       `scale(PMI)` = "PMI", 
                       `newStudyNIMH_HBCC` = "NIMH_HBCC",
                       `newStudyPitt` = "Pitt",
                       `newStudyMSSM` = "MSSM",
                       `newStudyPenn` = "Penn",
                       `scale(RIN)` = "RIN")) %>%
  merge(unique(acc_estimations_long[c('cell_class', 'cell_type')]), by.x = "cell_type")

#Repeat with cohort stratification
acc_lms_by_age_class = acc_estimations_long %>% 
  group_by(age_class, cell_type, newStudy) %>%
  do(tidy(lm(scale(rel_prop) ~ scale(RIN) + scale(PMI) +
             reportedGender + primaryDiagnosis, data = .))) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(`p.value`, method = 'fdr'),
         term = recode(term,
                       `(Intercept)` = "Intercept",
                       `reportedGendermale` = "gender:Male",
                       `primaryDiagnosisSchizophrenia` = "SCZ",
                       `scale(ageDeath)` = "Age",
                       `scale(PMI)` = "PMI", 
                       `newStudyNIMH_HBCC` = "NIMH_HBCC",
                       `newStudyPitt` = "Pitt",
                       `newStudyMSSM` = "MSSM",
                       `newStudyPenn` = "Penn",
                       `scale(RIN)` = "RIN")) %>%
  merge(unique(acc_estimations_long[c('cell_class', 'cell_type')]), by.x = "cell_type")


```

Model Microscopy CTPs
```{r}

micro_mega_lms = microCTPs_long %>% 
  group_by(cell_type) %>%
  do(tidy(lm(scale(rel_prop) ~ scale(PMI) + Subject.Group + Sex + scale(Age),  data = .))) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(`p.value`, method = 'fdr')) %>%
   mutate(term = recode(term,
                       `(Intercept)` = "Intercept",
                       `scale(PMI)` = "PMI", 
                       `Subject.GroupControl` = "Control",
                       `Subject.GroupSCHIZ`= "SCZ",
                       `SexM` = "Sex"))

```



------------////FIGURES////------------

Figure S1: Illustrating the association between cell prop and diagnosis and cohort specific differences
```{r, fig.height=8, fig.width=8 }
significant_combined = subset(acc_combined_lms, padj < 0.1 & cell_type %in% c("SST", "PVALB", "VIP") & term == "SCZ")
significant_combined$asterisks <- ifelse(significant_combined$padj <= 0.01, "***",
                                         ifelse(significant_combined$padj <= 0.05, "**", "*"))
significant_combined$vjust <- ifelse(significant_combined$estimate >= 0, -3, 4)

figure_S1b = acc_combined_lms %>%
  filter(term == 'SCZ', cell_type %in% c("SST", "PVALB", "VIP")) %>% 
  mutate(cell_type = fct_reorder(cell_type, estimate)) %>% 
  ggplot(aes(x = newStudy, y = estimate, fill = newStudy)) + 
  geom_hline(yintercept = 0) + 
  geom_bar(stat = "identity", position = "dodge") + 
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error)) + 
  ylab('Beta Coefficient') + 
  xlab('') +
  scale_fill_manual(values = cbPalette) +
  geom_text(data = significant_combined,
            aes(newStudy, estimate, label = asterisks), 
            vjust = significant_combined$vjust) +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(vjust = 6, size = 15),
    axis.title.y = element_text(size = 16),
    strip.text.x = element_text(size = 14),
    axis.line.x = element_blank(), 
    axis.ticks.x = element_blank()) +
  facet_grid(~cell_type, drop = T, scale = "free_x", space = "free") +
  guides(fill = guide_none()) +
  scale_y_continuous(limits = c(-1, 1))

# Figure 1c: Raw MGPs vs. Diagnosis
figure_S1c <- acc_estimations_long %>%
  filter(cell_type %in% c("SST", "PVALB", "VIP")) %>%
  ggplot(aes(x = primaryDiagnosis, y = rel_prop)) +
  geom_boxplot(outlier.shape = NA, aes(fill = newStudy, alpha = 0.5), show.legend = F, notch = T) +
  geom_beeswarm(size = 1, alpha = 0.3, aes(colour = newStudy), show.legend = F) +
  ylab('Relative Cell Prop.') + 
  xlab('Diagnosis') + 
  scale_x_discrete(labels = c("C", "SCZ")) +
  scale_colour_manual(values = cbPalette) +
  scale_fill_manual(values = cbPalette) +
  theme(
    axis.text.x = element_text(size = 15),
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(vjust = -2, size = 15),
    axis.title.y = element_text(size = 15),
    strip.text.x = element_text(size = 14),
    strip.text.y = element_text(size = 14)) +
  guides(fill = FALSE) +
  facet_grid(rows = vars(cell_type), cols = vars(newStudy), drop = T, scale = "free", switch = "y", space = "free") +
  geom_signif(comparisons = list(c("control", "Schizophrenia")), map_signif_level = TRUE, , test = wilcox.test) +
  scale_y_continuous(limits = c(-4, 4)) 

#Figure 1d - Hisotgram of cohort ages
figure_S1d = acc_estimations %>%
    ggplot(aes(x = ageDeath)) +
    geom_histogram(aes(fill = newStudy)) +
    facet_wrap(~ newStudy, nrow = 6, scales = "free_y") +
    labs(x = "ageDeath", y = "Frequency") +
    scale_fill_manual(values = c("#E69F00", "#D55E00", "#CC79A7")) + 
  scale_y_continuous(n.breaks = 4) +
  guides(fill = guide_none()) + 
  theme(panel.grid = element_blank(),
        strip.background = element_blank(),
        strip.text = element_text(size = rel(0.9), face = "plain"),
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.title.x = element_text( size = 15),
    axis.title.y = element_text(size = 16),
    strip.text.x = element_text(size = 14))

# Combine Figure 1
figure_S1_top = (figure_S1b | figure_S1d) + plot_layout(widths = c(2,1))
figure_S1_bottom = figure_S1c
  
figure_S1 = figure_S1_top / figure_S1_bottom + plot_annotation(tag_levels = 'A') + plot_layout(heights = c(2, 3)) & theme(plot.tag = element_text(size = 25))
figure_S1

#Save to plots folder
output_filename <- "~/cell_prop_psychiatry/plots/figure_S1.jpg"

# Use ggsave to save the plot as a JPG image
ggsave(output_filename, figure_S1, width = 10, height = 10, dpi = 200)

```



Figure 2: Introducing and presenting an interaction between age and diagnosis w/rt cell proportion.
```{r, fig.height=10, fig.width=10}
# Plot interaction beta values from meganalysis
significant_mega = subset(acc_mega_lms, padj <= 0.1 & term == "Age x Diagnosis")
significant_mega$asterisks = ifelse(significant_mega$padj <= 0.01, "***",
                                   ifelse(significant_mega$padj <= 0.05, "**", "*"))
significant_mega$vjust = ifelse(significant_mega$estimate >= 0, -2, 3)

# Figure S2a: Results from interaction meganalysis stratified by cell class - plot interaction beta coeff
acc_mega_lms$cell_type = factor(acc_mega_lms$cell_type, 
                                levels = c("PVALB", "SST", "VIP", "LAMP5", "PAX6", 
                                           "IT", "L4.IT", "L5.ET", "L5.6.IT.Car3", 
                                           "L5.6.NP", "L6.CT", "L6b", "Astrocyte", 
                                           "Endothelial", "Microglia", "Oligodendrocyte", 
                                           "OPC", "Pericyte", "VLMC"))

figure_S2a = acc_mega_lms %>% 
  filter(term %in% 'Age x Diagnosis') %>% 
  ggplot(aes(x = cell_type, y = estimate)) + 
  geom_hline(yintercept = 0) + 
  geom_bar(stat = "identity", position = "dodge") + 
  facet_grid(~cell_class, drop = T, scale = "free_x", space = "free") +
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error)) + 
  ylab('Interaction Coefficient') + 
  xlab('Cell Type') + 
  geom_text(data = significant_mega,
            aes(cell_type, estimate, label = asterisks), 
            vjust = significant_mega$vjust) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 12), 
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(vjust = 8, size = 15),
    axis.title.y = element_text(size = 15),
    strip.text.x = element_text(size = 14)) +
  guides(fill = FALSE) +
  scale_y_continuous(limits = c(-0.5, 0.7))

# Figure S2b: Plot beta values from meganalysis in 10-year age bins for SST, PVALB, VIP
significant_mega_age = subset(acc_lms_by_age, padj <= 0.1 & term == "SCZ" & cell_type %in% c("SST", "PVALB", "VIP"))
significant_mega_age$asterisks = ifelse(significant_mega_age$padj <= 0.01, "***",
                                         ifelse(significant_mega_age$padj <= 0.05, "**", "*"))
significant_mega_age$vjust = ifelse(significant_mega_age$estimate >= 0, -3, 3)

figure_S2b = acc_lms_by_age %>% 
  filter(term %in% 'SCZ' & cell_type %in% c("SST", "PVALB", "VIP")) %>% 
  mutate(cell_type = factor(cell_type, levels = c("PVALB", "SST", "VIP"))) %>% 
  ggplot(aes(x = age_range, y = estimate)) + 
  geom_hline(yintercept = 0) + 
  geom_bar(stat = "identity", position = "dodge") + 
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error)) + 
  ylab('Beta Coefficient') + 
  xlab('\n \n Age Range') + 
  geom_text(data = significant_mega_age,
            aes(age_range, estimate, label = asterisks), 
            vjust = significant_mega_age$vjust) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 13), 
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(vjust = 8, size = 15),
    axis.title.y = element_text(size = 15),
    strip.text.x = element_text(size = 14)) +
  facet_wrap(~cell_type, drop = T, scale = "free_x") +
  guides(fill = FALSE) +
  scale_y_continuous(limits = c(-1.4, 1.4))

# Figure S2c: Visualizing interaction between age and diagnosis by plotting rel_prop vs. age in each diagnosis 
figure_S2c = acc_estimations_long %>%
  filter(cell_type %in% c("SST", "PVALB", "VIP")) %>%
  ggplot(aes(x = ageDeath, y = rel_prop, color = primaryDiagnosis)) +
  geom_point(alpha = 0.5, size = 0.5) + 
  geom_smooth(se = F, method = 'lm', fullrange = T) +
  ylab('Relative Cell Prop.') + 
  xlab('Age') + 
  theme(
    axis.text.x = element_text(size = 15),
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(vjust = -1, size = 15),
    axis.title.y = element_text(size = 15),
    legend.position = c(0.9, 0.8),
    legend.title = element_blank(),
    legend.text = element_text(size = 12),  
    strip.text.x = element_text(size = 14)) +
  facet_grid(~cell_type, drop = T, scale = "free_x", space = "free") +
  guides(fill = guide_legend(nrow = 1)) +
  scale_y_continuous(limits = c(-2, 2)) + 
  scale_color_manual(values = c("dodgerblue2", "firebrick2"), labels=c('CON', 'SCZ'))

# Figure S2d: Mega-analysis results when binned as +/- 70 years
significant_mega_age_class = subset(acc_mega_lms_by_age_class, padj <= 0.1 & term == "SCZ" & cell_type %in% c("SST", "PVALB", "VIP"))
significant_mega_age_class$asterisks = ifelse(significant_mega_age_class$padj <= 0.01, "***", ifelse(significant_mega_age_class$padj <= 0.05, "**", "*"))
significant_mega_age_class$vjust = ifelse(significant_mega_age_class$estimate >= 0, -2, 3)

figure_S2d = acc_mega_lms_by_age_class %>% 
  filter(term %in% 'SCZ' & cell_type %in% c("SST", "PVALB", "VIP")) %>% 
  mutate(cell_type = factor(cell_type, levels = c("PVALB", "SST", "VIP"))) %>% 
  ggplot(aes(x = age_class, y = estimate, fill = age_class)) + 
  geom_hline(yintercept = 0) + 
  geom_bar(stat = "identity", position = "dodge") + 
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error)) + 
  ylab('Beta Coefficient') + 
  xlab('\n \n Age Range') + 
  scale_fill_manual(values = c("mediumpurple3", "olivedrab3"), labels=c('Above 70', 'Below 70')) +
  geom_text(data = significant_mega_age_class,
            aes(age_class, estimate, label = asterisks), 
            vjust = significant_mega_age_class$vjust) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 13), 
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(vjust = 8, size = 15),
    axis.title.y = element_text(size = 15),
    strip.text.x = element_text(size = 14)) +
  facet_wrap(~cell_type, drop = T, scale = "free_x") +
  guides(fill = FALSE) +
  scale_y_continuous(limits = c(-1, 1))

figure_S2_top = (figure_S2b + figure_S2d) + plot_layout(widths = c(1.6, 1))
figure_S2 = figure_S2_top / figure_S2c / figure_S2a
figure_S2 = figure_S2 + plot_annotation(tag_levels = 'A') + plot_layout(heights = c(1, 1, 1)) & theme(plot.tag = element_text(size = 25))
figure_S2


# Save to plots folder
output_filename <- "~/cell_prop_psychiatry/plots/figure_S2.jpg"

# Use ggsave to save the plot as a JPG image
ggsave(output_filename, figure_S2, width = 11, height = 11, dpi = 200)


```

Figure S3: Validating trends and interactions in ssRNAseq 
```{r, fig.height= 5, fig.width=9}
# Figure 3a: Relative cell proportions from snCTPs
figure_S3a <- microCTPs_long %>% 
  filter(Subject.Group %in% c('SCHIZ', "Control"), cell_type %in% c("SST", "PVALB", "VIP")) %>%
  ggplot(aes(x = Age_norm, y = rel_prop, color = Subject.Group)) +
  geom_point(alpha = 0.8, size = 1.3) + 
  geom_smooth(se = F, method = 'lm') +
  scale_colour_hue(labels = c("Control", "SCZ")) +
  ylab('Relative Cell Prop.') + 
  xlab('Age') + 
  theme(
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.title.x = element_text(vjust = -2, size = 16),
    axis.title.y = element_text(size = 16),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 12), 
    strip.text.x = element_text(size = 15),
    strip.text.y = element_text(size = 15)
  ) +
  scale_color_manual(values = c("dodgerblue2", "firebrick2"), labels=c('CON', 'SCZ')) +
  facet_wrap(~cell_type, drop = T, scale = "free_y")

# Figure 3b: microCTP Beta plot
significant_micro_mega <- subset(micro_mega_lms, padj <= 0.1 & term == "SCZ" & cell_type %in% c("SST", "PVALB", "VIP"))
significant_micro_mega$asterisks <- ifelse(significant_micro_mega$padj <= 0.01, "***",
                                         ifelse(significant_micro_mega$padj <= 0.05, "**", "*"))
significant_micro_mega$vjust <- ifelse(significant_micro_mega$estimate >= 0, -5, 5)

figure_S3b <- micro_mega_lms %>% 
  filter(term %in% 'SCZ' & cell_type %in% c("SST", "PVALB", "VIP")) %>% 
  mutate(cell_type = fct_reorder(cell_type, estimate)) %>% 
  ggplot(aes(x = cell_type, y = estimate)) + 
  geom_hline(yintercept = 0) + 
  geom_bar(stat = "identity", position = "dodge") + 
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error)) + 
  ylab('Beta Coefficient') + 
  xlab('Cell Type') + 
  geom_text(data = significant_micro_mega,
            aes(estimate, label = asterisks), 
            vjust = significant_micro_mega$vjust) +
  theme(
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 12), 
    strip.text.x = element_text(size = 15)
  ) 
#: Log transformation and scaling
microCTPs_long = microCTPs_long %>%
  mutate(Log_density = log(rel_prop + 1)) %>%
  group_by(cell_type) %>%
  filter(!Log_density %in% boxplot.stats(Log_density)$out) %>%
  ungroup() %>%
  drop_na()
#Residualize
microCTPs_long = microCTPs_long %>%
  mutate(resid = resid(lm(scale(Log_density) ~ scale(Age) + scale(PMI) +
               Sex, data = (microCTPs_long %>% group_by(cell_type))))) %>% ungroup()
microCTPs_long$Subject.Group = microCTPs_long$Subject.Group %>% factor(levels = c('Control', "SCHIZ"))

figure_S3c = microCTPs_long %>% filter(cell_type %in% c("PVALB", "SST", "VIP")) %>%
ggplot(aes(x = Subject.Group, y = resid, fill = Subject.Group)) +
geom_violin(aes(fill = Subject.Group), alpha = 0.4) +
  geom_quasirandom(na.rm = T, shape = 16, alpha = 0.9) + 
  stat_summary(fun = mean, geom = "point", aes(fill = Subject.Group), size = 4, shape = 23, col = "black") +
  ylab("Cellular density residuals\n(AU, ACC)") + 
    xlab('Diagnosis') + 
  scale_x_discrete(labels = c("CON", "SCZ")) +
  scale_colour_manual(values = c("dodgerblue2", "firebrick2")) +
  scale_fill_manual(values = c("dodgerblue2", "firebrick2")) +
  theme(
    axis.text.x = element_text(size = 15),
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(vjust = -2, size = 15),
    axis.title.y = element_text(size = 15),
    strip.text.x = element_text(size = 14),
    strip.text.y = element_text(size = 14)) +
  guides(fill = FALSE) +
  facet_wrap(~cell_type, ncol = 3, scales = "free_y") +
  geom_signif(comparisons = list(c('Control', "SCHIZ")), map_signif_level = F, test = wilcox.test)
  

# Figure 3: Combined figure
figure_S3 = (figure_S3c + figure_S3b) +
  theme(plot.margin = unit(c(5.5, 10, 5.5, 5.5), "pt"))
figure_S3 = figure_S3 + plot_annotation(tag_levels = 'A') + plot_layout(widths = c(3, 1)) & theme(plot.tag = element_text(size = 25))
figure_S3

#Save to plots folder
output_filename <- "~/cell_prop_psychiatry/plots/figure_S3.jpg"

# Use ggsave to save the plot as a JPG image
ggsave(output_filename, figure_S3, width = 9, height = 4, dpi = 200)

```


Figure S3c - Keon's Code  Version
```{r}

cell_dens_long = microCTPs_long %>%
  # log transformation for computing residuals 
  mutate(Log_density = log(rel_prop + 1)) %>%
  # drop NAs and outliers +/- 1.5 IQR   
  group_by(cell_type) %>%
  filter(!Log_density %in% boxplot.stats(Log_density)$out) %>% 
  ungroup() %>%
  drop_na()

# compute density residuals ----
# full model 
fit = lm(scale(Log_density) ~ Subject.Group + Sex + scale(Age) + scale(PMI), data = cell_dens_long)

# residualize and add DX back 
design = model.matrix(fit)
add_vars = grep("Subject.Group", colnames(design), value = T)
resid = residuals(fit) + fit$coefficients[add_vars] %*% t(design[,add_vars])
cell_dens_long$res = as.vector(resid)

p1 = cell_dens_long %>% 
  filter(cell_type %in% c("PVALB","SST","VIP")) %>%
  mutate(cell_type = factor(cell_type, levels = c("PVALB","SST","VIP"))) %>%
ggplot(aes(x = Subject.Group, y = res, color = Subject.Group)) +
  geom_violin(aes(fill = Subject.Group), alpha = 0.4) +
  geom_quasirandom(na.rm = T, shape = 16, alpha = 0.9) + 
  stat_summary(fun = mean, geom = "point", aes(fill = Subject.Group), size = 4, shape = 23, col = "black") +
  labs(y = 'Cellular density residuals\n(AU, ACC)', x = "", title = "", fill = "Diagnostic group", color = "Diagnostic group") +  
  scale_x_discrete(labels = c("CON", "SCZ")) +
  scale_colour_manual(values = c("dodgerblue2", "firebrick2")) +
  scale_fill_manual(values = c("dodgerblue2", "firebrick2")) +
  theme_classic2() +
  theme(plot.title = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        axis.text = element_text(colour = "black"),
        axis.line = element_line(),
        strip.text = element_text(face = "bold"),
        plot.margin = margin(0,0,-1,0,"cm")) +
  facet_wrap(~cell_type, ncol = 3, scales = "free_y") +
  geom_signif(comparisons = list(c('Control', "SCHIZ")), 
            map_signif_level = TRUE, 
            test = wilcox.test, 
            textsize = 3, 
            color = "black")
p1
```



```{r}
MGPVolcano <- function(matrix, marker_genes) {
  significant_genes <- subset(matrix, abs(logFC) > 1 & P.Value < 0.1)
  significant_marker_genes <- subset(matrix, gene_symbol %in% marker_genes & abs(logFC) > 1 & P.Value < 0.1)
  
  # Increase plot limits slightly to provide more space for labels
  x_range <- range(matrix$logFC, na.rm = TRUE)
  y_range <- range(-log10(matrix$P.Value), na.rm = TRUE)
  
  matrix %>%
    ggplot(aes(x = logFC, y = -log10(P.Value))) + 
    geom_vline(xintercept = -1, linetype = "dotted") +
    geom_vline(xintercept = 1, linetype = "dotted") +
    geom_hline(yintercept = -log10(0.1), linetype = "dotted") +
    geom_point(
      aes(label = gene_symbol), 
      alpha = 0.3, # Set opacity 30%
      color = "black", # Default color for non-significant points
      size = 3 # Adjust size as needed
    ) +
    geom_point(
      data = significant_marker_genes, 
      color = 'blue', # Set specific color for significant marker genes
      size = 3 # Keep size consistent
    ) +
    geom_point(
      data = subset(matrix, gene_symbol %in% marker_genes & !(gene_symbol %in% significant_marker_genes$gene_symbol)),
      color = 'red', # Set specific color for other specified marker genes
      size = 3 # Keep size consistent
    ) +
    geom_label_repel(
      data = significant_marker_genes, 
      aes(label = gene_symbol), 
      box.padding = 0.35, 
      point.padding = 0.5,
      segment.color = 'grey50',
      fill = "white", 
      color = "blue",
      max.overlaps = Inf,
      nudge_x = (x_range[2] - x_range[1]) * 0.2, # Nudge labels horizontally
      nudge_y = (y_range[2] - y_range[1]) * 0.2 # Nudge labels vertically
    ) +
    geom_label_repel(
      data = subset(matrix, gene_symbol %in% marker_genes & !(gene_symbol %in% significant_marker_genes$gene_symbol)), 
      aes(label = gene_symbol), 
      box.padding = 0.35, 
      point.padding = 0.5,
      segment.color = 'grey50',
      fill = "white", 
      color = "red",
 max.overlaps = Inf,
      nudge_x = (x_range[2] - x_range[1]) * 0.1, # Nudge labels horizontally
      nudge_y = (y_range[2] - y_range[1]) * 0.1 # Nudge labels vertically
    ) +
    theme_cowplot() +
    theme(legend.position = "none") +
    expand_limits(x = x_range + c(-1, 2) * 0.5, y = y_range + c(-1, 1) * 0.5) # Expand plot limits
}


```

Figure S5: LCMseq expression

```{r, fig.width= 6, fig.height=10}
scz_lcm_seq_de$genes = scz_lcm_seq_de$gene_symbol

LCMForestPlot <- function(data, marker_genes, min_logFC, max_logFC) {
  marker_data <- data[data$genes %in% marker_genes, ]
  marker_data <- marker_data %>%
    arrange(factor(genes, levels = marker_genes)) %>%
    mutate(y_position = as.numeric(factor(genes, levels = unique(genes))))
  
  label_positions <- unique(marker_data$genes)
  breaks_y <- sapply(label_positions, function(gene) {
    mean(marker_data$y_position[marker_data$genes == gene])
  })
  
  p <- ggplot(marker_data, aes(x = logFC, y = y_position, xmin = CI.L, xmax = CI.R)) +
    geom_point(size = 4) +
    geom_errorbarh(aes(height = 0.4), size = 1, width = 0.2) +
    geom_vline(xintercept = 0, linetype = "dotted") +
    xlab("Log Fold Change") + ylab("Gene") +
    scale_y_continuous(breaks = breaks_y, labels = label_positions) +
    coord_cartesian(xlim = c(min_logFC, max_logFC)) +
    theme(
      axis.title = element_text(size = 16),
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 14),
      legend.title = element_blank(),
      legend.text = element_text(size = 12),
      plot.title = element_text(size = 20, face = "bold"),
      strip.text = element_text(size = 16)
    )
  
  return(p)
}

PVALB_data = scz_lcm_seq_de %>% filter(CT == "PVALB")
SST_data = scz_lcm_seq_de %>% filter(CT == "SST")
VIP_data = scz_lcm_seq_de %>% filter(CT == "VIP")

PVALB_LCM_forest = LCMForestPlot(PVALB_data, PVALB_markers, -2.7, 1.5)
SST_LCM_forest = LCMForestPlot(SST_data, SST_markers, -1.4, 3)
VIP_LCM_forest = LCMForestPlot(VIP_data, VIP_markers, -2.2, 1.9)

combined_lcm_forest = ggarrange(PVALB_LCM_forest, SST_LCM_forest, VIP_LCM_forest,
                      nrow = 3,
                      hjust = 0,
                      heights = c(6, 5, 8),
                      common.legend = TRUE,
                      align = 'v',
                      font.label = list(size = 32, face = "bold"))
combined_lcm_forest

ggsave("~/cell_prop_psychiatry/plots/figure_S5.jpg", combined_lcm_forest, width = 6, height = 10, dpi = 400)
```

Figure S6: AgeOnset vs. Duration of Illness 
```{r, fig.width=9, fig.height=4}

# Filter and model data to residualize for ageOnset
ageOnset_model <- mgp_estimations_long %>% 
  filter(primaryDiagnosis %in% c('Schizophrenia'), cell_type %in% c("SST", "PVALB", "VIP"), dataset %in% c("GVEX", "LIBD_szControl")) %>% 
  group_by(dataset, cell_type) %>%
  do(tidy(lm(scale(rel_prop) ~ scale(ageDeath) + scale(RIN) + scale(PMI) + reportedGender + scale(ageOnset), data = .)))

# Initialize empty dataframe for residuals
CELL_TYPES <- c("SST", "PVALB", "VIP")
DATASETS <- c("GVEX", "LIBD_szControl")
ageOnset_residuals_data <- matrix(ncol = (ncol(mgp_estimations_long) + 1), nrow = 0) %>% data.frame()
colnames(ageOnset_residuals_data) <- c(colnames(mgp_estimations_long), "resid")

# Calculate residuals
for (DATASET in DATASETS) {
  for (CELL_TYPE in CELL_TYPES) {
    res_model_data <- mgp_estimations_long %>% filter(primaryDiagnosis %in% c('Schizophrenia'), cell_type == CELL_TYPE, dataset == DATASET)
    res_model <- lm(scale(rel_prop) ~ scale(ageDeath) + scale(RIN) + scale(PMI) + reportedGender, data = res_model_data)
    res_model_data <- res_model_data %>% add_residuals(var = "resid", model = res_model)
    ageOnset_residuals_data <- ageOnset_residuals_data %>% rbind(res_model_data)
  }
}


figure_S5 <- ageOnset_residuals_data %>%
  ggplot(aes(x = ageOnset, y = durationIllness, color = dataset)) +
  geom_point(alpha = 0.5, size = 0.5) + 
  geom_smooth(se = T, fullrange = TRUE, method = "lm") +
  ylab('Duration of SCZ (years)') + 
  xlab('Age of SCZ onset (years)') + 
  theme(
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(vjust = -2, size = 16, margin = margin(t = 8)),
    axis.title.y = element_text(size = 16),
    strip.text = element_text(size = 16),
    legend.text = element_text(size = 16),
    legend.position = "none", 
    plot.margin = margin(10, 10, 20, 10)) +
  facet_wrap(~ dataset) +
  scale_color_manual(values = c("dodgerblue2", "firebrick2")) +
  guides(color = guide_legend(title = NULL)) +
  stat_cor(aes(label = paste(..r.label.., ..p.label.., sep = "~")), color = "black", geom = "label")


figure_S5
```

Figure S6: rCTPs vs Duration of Illness
```{r, fig.width=9, fig.height=6}
figure_S6 <- ageOnset_residuals_data %>%
  ggplot(aes(x = durationIllness, y = resid, color = dataset)) +
  geom_point(alpha = 0.5, size = 0.5) + 
  geom_smooth(se = FALSE, method = 'lm', fullrange = TRUE) +
  ylab('rCTP residuals (AU)') + 
  xlab('Duration of SCZ (years)') + 
  theme(
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(vjust = -2, size = 16, margin = margin(t = 8)),
    axis.title.y = element_text(size = 16),
    strip.text = element_text(size = 16),
    legend.text = element_text(size = 16),
    legend.position = "none", 
    plot.margin = margin(10, 10, 20, 10)) +
  facet_grid(cols = vars(cell_type), rows = vars(dataset)) +
  scale_y_continuous(limits = c(-1.5, 1.5)) + 
  scale_color_manual(values = c("dodgerblue2", "firebrick2")) +
  guides(color = guide_legend(title = NULL)) +
  stat_cor(aes(label = paste(..r.label.., ..p.label.., sep = "~")), color = "black", geom = "label")

figure_S6
```

figure S7: mean cohort age vs. rctp beta
```{r, fig.width=5, fig.height=8}
cbPalette = c("#56B4E9", "#009E73","#E69F00", "#0072B2", "#D55E00", "#CC79A7","#000000","#F0E442")

cohort_means = mgp_estimations %>%
  group_by(newStudy) %>%
  summarize(mean_ageDeath = mean(ageDeath, na.rm = TRUE))

data = combined_lms %>% filter(cell_type %in% c("PVALB", "SST", "VIP"), term == "SCZ") %>% left_join(cohort_means, by = "newStudy")

figure_s7 =  data %>% 
  ggplot(aes(x = mean_ageDeath, y = estimate)) +
  geom_point(aes(color = newStudy), alpha = 0.5, size = 2) + 
  geom_smooth(se = TRUE, method = 'lm', fullrange = TRUE, color = "blue") +
  geom_hline(yintercept = 0, linetype = "dotted", color = "black") + 
  scale_color_manual(values = cbPalette) + 

  ylab('SCZ rCTP diff. (std. beta)') + 
  xlab('Mean age at death (years)') + 
  theme(
    axis.text.x = element_text(size = 15),
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(vjust = -2, size = 15),
    axis.title.y = element_text(size = 15),
    strip.text.x = element_text(size = 14),
    legend.position = "none"
  ) +
  facet_wrap(~cell_type, ncol = 1, scales = "free_y") +
  stat_cor(aes(label = paste(..r.label.., ..p.label.., sep = "~")), color = "black")

  figure_s7

```

Figure S8: LCMseq cell densitiess
```{r}

```

