---
title: "Untitled"
author: "Daniel Kiss"
date: "11/6/2023"
output: html_document
---


```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(Seurat)
  library(edgeR)
  library(ggrepel)
  library(cowplot)
  library(limma)
  library(ggpubr)
  library(gprofiler2)
  library(EnhancedVolcano)
  library(patchwork)
})
theme_set(theme_classic2())
#Colour palette
cbPalette <- c("#56B4E9", "#009E73","#E69F00", "#0072B2", "#D55E00", "#CC79A7","#000000","#F0E442")
```


Load Data // Define Variables
```{r}
pseudobulk = read.delim("/external/rprshnas01/netdata_kcni/stlab/SZBDMulticohort/pseudobulk_fine_grained.counts.tsv")
pseudobulk_metadata = read.delim("/external/rprshnas01/netdata_kcni/stlab/SZBDMulticohort/pseudobulk_fine_grained.metadata.tsv")
CELL_TYPES = pseudobulk_metadata$Celltype %>% unique()
```

Define Subjects
```{r}
USE_COHORT = 'MtSinai'
NUM_CELLS = 20

SST_SCZ = pseudobulk_metadata %>% filter(Celltype == "In-SST", Cohort == USE_COHORT, num_cells > NUM_CELLS, Phenotype == "SZ") %>% pull(unique_donor_ID)
SST_CON = pseudobulk_metadata %>% filter(Celltype == "In-SST", Cohort == USE_COHORT, num_cells > NUM_CELLS, Phenotype == "CON") %>% pull(unique_donor_ID)
PVALB_SCZ = pseudobulk_metadata %>% filter(Celltype == "In-PV_Basket", Cohort == USE_COHORT, num_cells > NUM_CELLS, Phenotype == "SZ") %>% pull(unique_donor_ID)
PVALB_CON = pseudobulk_metadata %>% filter(Celltype == "In-PV_Basket", Cohort == USE_COHORT, num_cells > NUM_CELLS, Phenotype == "CON") %>% pull(unique_donor_ID)

```

Subset count matrices and convert to cpm
```{r}
sst_scz_counts = pseudobulk %>% filter(unique_donor_ID %in% SST_SCZ,  Celltype == "In-SST") 
sst_scz_counts = sst_scz_counts[, 3:ncol(sst_scz_counts)] %>% t() %>% cpm() %>% t() %>% as.data.frame()
rownames(sst_scz_counts) = SST_SCZ
  
sst_con_counts = pseudobulk %>% filter(unique_donor_ID %in% SST_CON,  Celltype == "In-SST") 
sst_con_counts = sst_con_counts[, 3:ncol(sst_con_counts)] %>% t() %>% cpm() %>% t() %>% as.data.frame()
rownames(sst_con_counts) = SST_CON
  
pvalb_scz_counts = pseudobulk %>% filter(unique_donor_ID %in% PVALB_SCZ,  Celltype == "In-PV_Basket") 
pvalb_scz_counts = pvalb_scz_counts[, 3:ncol(pvalb_scz_counts)] %>% t()%>% cpm() %>% t() %>% as.data.frame()
rownames(pvalb_scz_counts) = PVALB_SCZ
  
pvalb_con_counts = pseudobulk %>% filter(unique_donor_ID %in% PVALB_CON, Celltype == "In-PV_Basket")
pvalb_con_counts = pvalb_con_counts[, 3:ncol(pvalb_con_counts)] %>% t() %>% cpm() %>% t() %>% as.data.frame()
rownames(pvalb_con_counts) = PVALB_CON
```

Create a data frame with expression data for SST and PVALB mRNA
```{r}
counts_for_plot = data.frame(
  cell_type = rep(c("In-SST", "In-PVALB"),
                  c((length(SST_SCZ) + length(SST_CON)), 
                    (length(PVALB_SCZ) + length(PVALB_CON)))),
  phenotype = rep(c("SCZ", "CON", "SCZ", "CON"),
                  c(length(SST_SCZ), length(SST_CON), length(PVALB_SCZ), length(PVALB_CON))),
  cpm = c(sst_scz_counts$SST, sst_con_counts$SST, pvalb_scz_counts$PVALB, pvalb_con_counts$PVALB)
)
```

Plot data from above
```{r}
p1 = counts_for_plot %>% filter(cell_type == "In-PVALB") %>%
  ggplot(aes(x = phenotype, y = cpm)) +
  geom_boxplot(aes(alpha = 0.5), show.legend = F) + 
  facet_grid(cols = vars(cell_type)) +
  geom_signif(comparisons = list(c("CON", "SCZ")), map_signif_level = TRUE)

p2 = counts_for_plot %>% filter(cell_type == "In-SST") %>%
  ggplot(aes(x = phenotype, y = cpm)) +
  geom_boxplot(aes(alpha = 0.5), show.legend = F) + 
  facet_grid(cols = vars(cell_type)) +
  geom_signif(comparisons = list(c("CON", "SCZ")), map_signif_level = TRUE)

p1 +p2

```


figure_1c <- all_estimations_long %>%
  filter(primaryDiagnosis %in% c('Schizophrenia', "control"), cell_type %in% c("Inh_SST", "Inh_PVALB", "Inh_VIP")) %>%
  ggplot(aes(x = primaryDiagnosis, y = rel_prop)) +
  geom_boxplot(outlier.shape = NA, aes(fill = newStudy, alpha = 0.5), show.legend = F, notch = T) +
  geom_beeswarm(size = 1, alpha = 0.3, aes(colour = newStudy), show.legend = F) +
  ylab('Relative Cell Prop.') + 
  xlab('Diagnosis') + 
  scale_x_discrete(labels = c("C", "SCZ")) +
  scale_colour_manual(values = cbPalette) +
  scale_fill_manual(values = cbPalette) +
  theme(
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.title.x = element_text(vjust = -2, size = 16),
    axis.title.y = element_text(size = 16),
    strip.text.x = element_text(size = 14)) +
  guides(fill = FALSE) +
  facet_grid(rows = vars(cell_type), cols = vars(newStudy), drop = T, scale = "free", switch = "y", space = "free") +
  geom_signif(comparisons = list(c("control", "Schizophrenia")), map_signif_level = TRUE) +
  scale_y_continuous(limits = c(-3.5, 3.5)) 


