---
title: "psychencode_DE_figure"
author: "Daniel Kiss"
date: "1/16/2024"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(Seurat)
  library(edgeR)
  library(ggrepel)
  library(cowplot)
  library(limma)
  library(ggpubr)
  library(gprofiler2)
  library(EnhancedVolcano)
  library(patchwork)
})
theme_set(theme_classic2())
#Colour palette
cbPalette <- c("#56B4E9", "#009E73","#E69F00", "#0072B2", "#D55E00", "#CC79A7","#000000","#F0E442")
```

Load Data // Define Variables
```{r}
DE_results_ALL = readRDS("~/cell_prop_psychiatry/data/DE_results_ALL")
DE_results_McLean = readRDS("~/cell_prop_psychiatry/data/DE_results_McLean")
DE_results_MtSinai = readRDS("~/cell_prop_psychiatry/data/DE_results_MtSinai")
pseudobulk = read.csv("~/cell_prop_psychiatry/data/psychencode_label_transferred_pseudobulk_raw.csv")[,-1]
pseudobulk_metadata = read.csv("~/cell_prop_psychiatry/data/psychencode_label_transferred_pseudobulk_metadata.csv")[,-1] %>%
  group_by(ID) %>%
  mutate(total_cells = sum(num_cells)) %>%
  ungroup() 
#Add log10 number of cells variable
pseudobulk_metadata$log10_num_cells = log10(pseudobulk_metadata$num_cells)
#Add age group variable
pseudobulk_metadata =  pseudobulk_metadata %>%
  mutate(age_group = case_when(
    Age < 70 ~ "Under_70", Age >= 70 ~ "Over_70")) 
pseudobulk_metadata$age_group = pseudobulk_metadata$age_group %>% factor(levels = c("Under_70", "Over_70"))
pseudobulk_combined = left_join(pseudobulk_metadata, pseudobulk)

#Convert to log2(cmp + 1) / num_cells
genes = pseudobulk_combined[, 29:ncol(pseudobulk_combined)] %>% cpm(log=TRUE) %>% as.data.frame()

pseudobulk_transformed = cbind(pseudobulk_combined[,1:28], genes)
```

Important Variables to Define
```{r}
CELL_TYPES = c("SST", "PVALB", "VIP")
NUM_CELLS = 1000
AGE_THRESH = 70 
DATASET = c("MtSinai") #Vector of dataset names (can be length = 1)
```


Create a list of relevant data for each cell type
```{r}
no_batch_list = list()
#Subset count matrix (if doing cohort stratified analysis)
pseudobulk_subset = pseudobulk_transformed %>% filter(Cohort %in% DATASET)

#Loop through ALL cell types and perform above analyses
for(CELL_TYPE in CELL_TYPES) {
  counts_under_70 = pseudobulk_subset %>% filter(Celltype == CELL_TYPE, age_group == "Under_70", total_cells >= NUM_CELLS)
  counts_over_70 = pseudobulk_subset %>% filter(Celltype == CELL_TYPE, age_group == "Over_70", total_cells >= NUM_CELLS)

  # Create batch and covariate labels for Under_70
  disease = factor(counts_under_70 %>% pull(Phenotype), levels = c('CON', 'SZ'))
  sex = counts_under_70 %>% pull(Gender)
  pmi = counts_under_70 %>% pull(PMI)
  age = counts_under_70 %>% pull(Age)
  cohort = if (length(unique(DATASET)) > 1) counts_under_70 %>% pull(Cohort) else NULL
  log_10_cells_per_donor = counts_under_70 %>% pull(log10_num_cells)
  
  # Use variables to regress out covariates using removeBatchEffect
  data = counts_under_70[, 31:ncol(counts_under_70)] %>% t()
  under_70_no_batch = removeBatchEffect(x = data,
                                        batch = sex,
                                        batch2 = if (!is.null(cohort)) cohort else NULL,
                                        covariates = cbind(pmi, age, log_10_cells_per_donor),
                                        design = model.matrix(~ disease)) %>% t()
  under_70_no_batch = cbind(counts_under_70[, 1:30], under_70_no_batch)
  
  # Create batch and covariate labels for Over_70
  disease = factor(counts_over_70 %>% pull(Phenotype), levels = c('CON', 'SZ'))
  sex = counts_over_70 %>% pull(Gender)
  pmi = counts_over_70 %>% pull(PMI)
  age = counts_over_70 %>% pull(Age)
  cohort = if (length(unique(DATASET)) > 1) counts_over_70 %>% pull(Cohort) else NULL
  log_10_cells_per_donor = counts_over_70 %>% pull(log10_num_cells)
  
  # Use variables to regress out covariates using removeBatchEffect
  data = counts_over_70[, 31:ncol(counts_over_70)] %>% t()
  over_70_no_batch = removeBatchEffect(x = data,
                                       batch = sex,
                                       batch2 = if (!is.null(cohort)) cohort else NULL,
                                       covariates = cbind(pmi, age, log_10_cells_per_donor),
                                       design = model.matrix(~ disease)) %>% t()
  over_70_no_batch = cbind(counts_over_70[, 1:30], over_70_no_batch)
  
  #Create lists of data for each group
  joined_data = rbind(under_70_no_batch, over_70_no_batch) 
  no_batch_list[[CELL_TYPE]] = joined_data
}

```


Plot mRNA expression for main interneuron subtypes and their corresponding marker genes
```{r, fig.height= 12, fig.width= 7}
#Create separate plots for each cell type
SST_max_y <- max(no_batch_list$SST$SST, na.rm = TRUE) + 1  # Find max y-value and add 1 for space
SST_mRNA <- no_batch_list$SST %>%
  ggplot(aes(x = Phenotype, y = SST)) +
  geom_boxplot(outlier.shape = NA, show.legend = F) +
  facet_grid(~age_group, drop = T, scales = "free_y") +
  theme(text = element_text(size = 18)) +
  ylab('log2(cpm + 1)') + 
  geom_signif(comparisons = list(c("CON", "SZ")), map_signif_level = TRUE, y_position = SST_max_y - 1) +
  coord_cartesian(ylim = c(NA, SST_max_y))  # Adjust the y-axis limit

PVALB_max_y <- max(no_batch_list$PVALB$PVALB, na.rm = TRUE) + 1
PVALB_mRNA <- no_batch_list$PVALB %>% 
  ggplot(aes(x = Phenotype, y = PVALB)) +
  geom_boxplot(outlier.shape = NA, show.legend = F) +
  facet_grid(~age_group, drop = T, scales = "free_y") +
  theme(text = element_text(size = 18)) +
  ylab('log2(cpm + 1)') + 
  geom_signif(comparisons = list(c("CON", "SZ")), map_signif_level = TRUE, y_position = PVALB_max_y - 1) +
  coord_cartesian(ylim = c(NA, PVALB_max_y))

VIP_max_y <- max(no_batch_list$VIP$VIP, na.rm = TRUE) + 1
VIP_mRNA <- no_batch_list$VIP %>%
  ggplot(aes(x = Phenotype, y = VIP)) +
  geom_boxplot(outlier.shape = NA, show.legend = F) +
  facet_grid(~age_group, drop = T, scales = "free_y") +
  theme(text = element_text(size = 18)) +
  ylab('log2(cpm + 1)') + 
  geom_signif(comparisons = list(c("CON", "SZ")), map_signif_level = TRUE, y_position = VIP_max_y - 1) +
  coord_cartesian(ylim = c(NA, VIP_max_y))


mRNA_figure = SST_mRNA / PVALB_mRNA / VIP_mRNA + 
  plot_annotation(tag_levels = 'A') + 
  plot_layout(heights = c(1, 1, 1)) & 
  theme(plot.tag = element_text(size = 25))

```


Create volcano plots from DE results
```{r}
#Create vectors of MGP marker genes for each cell type
markers = read.csv("~/cell_prop_psychiatry/data/MGP_markers.csv")
SST_markers = markers %>% filter(subclass == "SST") %>% pull(gene)
PVALB_markers = markers %>% filter(subclass == "PVALB") %>% pull(gene)
VIP_markers = markers %>% filter(subclass == "VIP") %>% pull(gene)
```

Function to create a volcano plot and highlight genes in a gene list:
-Takes gene expression matrix and gene list
-Makes volcano plot
-Highlights non-significant marker genes in red and significant marker genes in blue
```{r}
MGPVolcano <- function(matrix, marker_genes) {
  significant_genes <- subset(matrix, abs(logFC) > 1 & P.Value < 0.1)
  significant_marker_genes <- subset(matrix, genes %in% marker_genes & abs(logFC) > 1 & P.Value < 0.1)
  
  # Increase plot limits slightly to provide more space for labels
  x_range <- range(matrix$logFC, na.rm = TRUE)
  y_range <- range(-log10(matrix$P.Value), na.rm = TRUE)
  
  matrix %>%
    ggplot(aes(x = logFC, y = -log10(P.Value))) + 
    geom_vline(xintercept = -1, linetype = "dotted") +
    geom_vline(xintercept = 1, linetype = "dotted") +
    geom_hline(yintercept = -log10(0.1), linetype = "dotted") +
    geom_point(
      aes(label = genes), 
      alpha = 0.3, # Set opacity 30%
      color = "black", # Default color for non-significant points
      size = 3 # Adjust size as needed
    ) +
    geom_point(
      data = significant_marker_genes, 
      aes(color = 'blue'), # Set specific color for significant marker genes
      size = 3 # Keep size consistent
    ) +
    geom_point(
      data = subset(matrix, genes %in% marker_genes & !(genes %in% significant_marker_genes$genes)),
      aes(color = 'red'), # Set specific color for other specified marker genes
      size = 3 # Keep size consistent
    ) +
    geom_label_repel(
      data = significant_marker_genes, 
      aes(label = genes), 
      box.padding = 0.35, 
      point.padding = 0.5,
      segment.color = 'grey50',
      fill = "white", 
      color = "blue",
      max.overlaps = Inf,
      nudge_x = (x_range[2] - x_range[1]) * 0.2, # Nudge labels horizontally
      nudge_y = (y_range[2] - y_range[1]) * 0.2 # Nudge labels vertically
    ) +
    geom_label_repel(
      data = subset(matrix, genes %in% marker_genes & !(genes %in% significant_marker_genes$genes)), 
      aes(label = genes), 
      box.padding = 0.35, 
      point.padding = 0.5,
      segment.color = 'grey50',
      fill = "white", 
      color = "red",
      max.overlaps = Inf,
      nudge_x = (x_range[2] - x_range[1]) * 0.1, # Nudge labels horizontally
      nudge_y = (y_range[2] - y_range[1]) * 0.1 # Nudge labels vertically
    ) +
    theme_cowplot() +
    theme(legend.position = "none") +
    expand_limits(x = x_range + c(-1, 2) * 0.5, y = y_range + c(-1, 1) * 0.5) # Expand plot limits
}

```

Create volcano plots for each cell for above and below age 70
```{r, fig.height= 16, fig.width= 12}
# Conditionally set DE_results basedon datasets  used
if (length(DATASET) > 1) {
  DE_results <- DE_results_ALL
} else {
  DE_results <- get(paste0("DE_results_", DATASET))
}

#Set figure font sizes
common_theme = theme(
  axis.title = element_text(size = 16),
  axis.text = element_text(size = 16)
)
#Make volcano plots
SST_volcano_under = MGPVolcano(DE_results$SST$DE_tables$DE_under, SST_markers) + ggtitle("Under 70") +
                    theme(plot.title = element_text(hjust = 0.5, size = 18)) + common_theme
SST_volcano_over = MGPVolcano(DE_results$SST$DE_tables$DE_over, SST_markers) + ggtitle("Over 70") +
                    theme(plot.title = element_text(hjust = 0.5, size = 18)) + common_theme

PVALB_volcano_under = MGPVolcano(DE_results$PVALB$DE_tables$DE_under, PVALB_markers) + ggtitle("Under 70") +
                    theme(plot.title = element_text(hjust = 0.5, size = 18)) + common_theme
PVALB_volcano_over = MGPVolcano(DE_results$PVALB$DE_tables$DE_over, PVALB_markers) + ggtitle("Over 70") +
                    theme(plot.title = element_text(hjust = 0.5, size = 18)) + common_theme

VIP_volcano_under = MGPVolcano(DE_results$VIP$DE_tables$DE_under, VIP_markers) + ggtitle("Under 70") +
                    theme(plot.title = element_text(hjust = 0.5, size = 18)) + common_theme
VIP_volcano_over = MGPVolcano(DE_results$VIP$DE_tables$DE_over, VIP_markers) + ggtitle("Over 70") +
                    theme(plot.title = element_text(hjust = 0.5, size = 18)) + common_theme


```

Combine into final DE_figure
```{r, fig.height= 18, fig.width= 16}
#Define plot title
dataset_title = if(length(DATASET) > 1) paste(DATASET, collapse = " & ") else DATASET

#Combine plots
SST_combined = SST_mRNA + SST_volcano_under + SST_volcano_over + plot_layout(widths = c(2,3,3)) + 
               plot_annotation(title = "SST", theme = theme(plot.title = element_text(hjust = 0.5, size = 22, face = "bold"), plot.margin = margin(t = 25, unit = "pt")))

PVALB_combined = PVALB_mRNA + PVALB_volcano_under + PVALB_volcano_over + plot_layout(widths = c(2,3,3)) +  
                 plot_annotation(title = "PVALB", theme = theme(plot.title = element_text(hjust = 0.5, size = 22, face = "bold"), plot.margin = margin(t = 25, unit = "pt")))

VIP_combined = VIP_mRNA + VIP_volcano_under + VIP_volcano_over + plot_layout(widths = c(2,3,3)) +  
               plot_annotation(title = "VIP", theme = theme(plot.title = element_text(hjust = 0.5, size = 22, face = "bold"), plot.margin = margin(t =25, unit = "pt")))

DE_figure = ggarrange(SST_combined, PVALB_combined, VIP_combined,
                      labels = c("A", "B", "C"),
                      ncol = 1, nrow = 3,
                      font.label = list(size = 32, face = "bold")) %>% 
                      annotate_figure(top = text_grob(paste("DE Analysis -", dataset_title), size = 28, face = "bold"))


#Save to plots folder
dataset_suffix = if(length(DATASET) > 1) "ALL" else DATASET
output_filename <- paste0("~/cell_prop_psychiatry/plots/DE_figure_", dataset_suffix ,".jpg")
ggsave(output_filename, DE_figure, width = 16, height = 18, dpi = 300)
```








