---
title: "psychencode_metadata_analyses"
author: "Daniel Kiss"
date: "9/26/2023"
output: html_document
---

The following code should be run based on the variables generated in lines 1-72 of psychencode_figures_cleaned.Rmd

Plot CTPs vs. ageOnset for each cohort
```{r}
#Note that CMC does not have ageOnset info - therefore we filter for dataset = GVEX and LIBD
ctp_vs_ageonset = all_estimations_long %>%
  filter(primaryDiagnosis %in% c('Schizophrenia'), cell_type %in% c("Inh_SST", "Inh_PVALB", "Inh_VIP"), dataset %in% c("GVEX", "LIBD")) %>%
  ggplot(aes(x = ageOnset, y = rel_prop, color = dataset)) +
  geom_point(alpha = 0.5, size = 0.5) + 
  geom_smooth(se = F, method = 'lm', fullrange = T) +
  ylab('Relative Cell Prop.') + 
  xlab('Age of Onset') + 
  theme(
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.title.x = element_text(vjust = -2, size = 16),
    axis.title.y = element_text(size = 16),
    strip.text.x = element_text(size = 16)) +
  facet_grid(~cell_type) +
  scale_y_continuous(limits = c(-2, 2))

ctp_vs_ageonset

#Now let's plot the difference between ageDeath and ageOnset to calculate diagnosisLength
ctp_vs_diagnosislength = all_estimations_long %>%
  filter(primaryDiagnosis %in% c('Schizophrenia'), cell_type %in% c("Inh_SST", "Inh_PVALB", "Inh_VIP"), dataset %in% c("GVEX", "LIBD")) %>%
  mutate(diagnosisLength = ageDeath - ageOnset) %>%
  ggplot(aes(x = diagnosisLength, y = rel_prop, color = dataset)) +
  geom_point(alpha = 0.5, size = 0.5) + 
  geom_smooth(se = F, method = 'lm', fullrange = T) +
  ylab('Relative Cell Prop.') + 
  xlab('Length of Diagnosis') + 
  theme(
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.title.x = element_text(vjust = -2, size = 16),
    axis.title.y = element_text(size = 16),
    strip.text.x = element_text(size = 16)) +
  facet_grid(~cell_type) +
  scale_y_continuous(limits = c(-2, 2))

ctp_vs_diagnosislength

#ageOnset vs. ageDeath
ageonset_vs_agedeath = all_estimations %>%
  filter(primaryDiagnosis %in% c('Schizophrenia'), dataset %in% c("GVEX", "LIBD")) %>%
  ggplot(aes(x = ageOnset, y = ageDeath, color = dataset)) +
  geom_point(alpha = 0.5, size = 0.5) + 
  geom_smooth(se = F, method = 'lm', fullrange = T) +
  ylab('Age of Death.') + 
  xlab('Age of Onset') + 
  theme(
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 13),
    axis.title.x = element_text(vjust = -2, size = 16),
    axis.title.y = element_text(size = 16),
    strip.text.x = element_text(size = 16))
  scale_y_continuous(limits = c(-2, 2))

ageonset_vs_agedeath
                                
```

Replot ageOnset vs. CTP but regress out ageDeath
```{r, fig.width=4, fig.height=3.5}
#Model for ageOnset
ageOnset_model = all_estimations_long %>% 
  filter(primaryDiagnosis %in% c('Schizophrenia'), cell_type %in% c("Inh_SST", "Inh_PVALB", "Inh_VIP"), dataset %in% c("GVEX", "LIBD")) %>% 
  group_by(dataset, cell_type) %>%
    do(tidy(lm(scale(rel_prop) ~ scale(ageDeath) + scale(RIN) + scale(PMI) + reportedGender +  scale(ageOnset), data = .)))

CELL_TYPES = c("Inh_SST", "Inh_PVALB", "Inh_VIP")
DATASETS = c("GVEX", "LIBD")
ageOnset_residuals_data = matrix(ncol = (ncol(all_estimations_long)+1), nrow = 0) %>% data.frame()
colnames(ageOnset_residuals_data) = c(colnames(all_estimations_long), "resid")

for(DATASET in DATASETS) {
  for(CELL_TYPE in CELL_TYPES) {
    #Create res_model without ageOnset variable
    res_model_data = all_estimations_long %>% filter(primaryDiagnosis %in% c('Schizophrenia'), cell_type == CELL_TYPE , dataset == DATASET)
    #Model without ageOnset
    res_model = lm(scale(rel_prop) ~ scale(ageDeath) + scale(RIN) + scale(PMI) + reportedGender, data = res_model_data)
    #Calculate residuals and add column to res_model_data
    res_model_data = res_model_data %>% add_residuals(var = "resid", model = res_model)
    ageOnset_residuals_data = ageOnset_residuals_data %>% rbind(res_model_data)
  }
}

#Plot ageOnst vs. Residuals
resid_vs_ageonset <- ageOnset_residuals_data %>%
  ggplot(aes(x = ageOnset, y = resid, color = dataset)) +
  geom_point(alpha = 0.5, size = 0.5) + 
  geom_smooth(se = FALSE, method = 'lm', fullrange = TRUE) +
  ylab('Residualized CTP Estimate') + 
  xlab('Age of Onset') + 
  theme(
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(vjust = -2, size = 16, margin = margin(t = 8)),
    axis.title.y = element_text(size = 16),
    strip.text = element_text(size = 16),
    legend.text = element_text(size = 16),
    plot.margin = margin(10, 10, 20, 10)) +
  facet_grid(cols = vars(cell_type), rows = vars(dataset)) +
  scale_y_continuous(limits = c(-1.5, 1.5)) + 
  scale_color_manual(values = c("dodgerblue2", "firebrick2")) +
  guides(color = guide_legend(title = NULL)) +
  stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~")), color = "black", geom = "label")


resid_vs_ageonset

#Save to plot folder
output_filename <- "~/cell_prop_psychiatry/plots/figure_4.pdf"

# Use ggsave to save the plot as a JPG image
ggsave(output_filename, resid_vs_ageonset, width = 8, height = 6, dpi = 200)

```

Let's look into correlations between cause of death and the residuals of PV and SST proportions.
```{r}
CELL_TYPES = c("Inh_SST", "Inh_PVALB", "Inh_VIP")
DATASETS = c("GVEX", "LIBD")
residuals_data = matrix(ncol = (ncol(all_estimations_long)+1), nrow = 0) %>% data.frame()
colnames(residuals_data) = c(colnames(all_estimations_long), "resid")

for(DATASET in DATASETS) {
  for(CELL_TYPE in CELL_TYPES) {
    #Create res_model 
    res_model_data = all_estimations_long %>% filter(primaryDiagnosis %in% c('Schizophrenia', 'control'), cell_type == CELL_TYPE , dataset == DATASET)
    res_model = lm(scale(rel_prop) ~ scale(ageDeath) + scale(RIN) + scale(PMI) + reportedGender, data = res_model_data)
    #Calculate residuals and add column to res_model_data
    res_model_data = res_model_data %>% add_residuals(var = "resid", model = res_model)
    residuals_data = residuals_data %>% rbind(res_model_data)
  }
}

#Add filtered causeDeath column
residuals_data  = residuals_data %>%
    mutate(causeDeath_filtered = ifelse(grepl("SUIC", toupper(causeDeath)), "suicide", "other"))

#Plot causeDeath vs. Residuals
resid_vs_ageonset = residuals_data %>%
  filter(primaryDiagnosis == 'Schizophrenia') %>%
  ggplot(aes(x = causeDeath_filtered, y = ageDeath)) +
  geom_boxplot() +
  ylab('Age of Death') + 
  xlab('Cause of Death') + 
  theme(
    axis.text.x = element_text(size = 13, angle = 90),
    axis.text.y = element_text(size = 13),
    axis.title.x = element_text(vjust = -2, size = 16),
    axis.title.y = element_text(size = 16),
    strip.text.x = element_text(size = 16)) +
  facet_grid(cols = vars(newStudy)) 
  #scale_y_continuous(limits = c(-2, 2)) 
resid_vs_ageonset

#Plot causeDeath vs. Residuals
resid_vs_ageonset = residuals_data %>% 
  filter(primaryDiagnosis == 'Schizophrenia') %>%
  ggplot(aes(x = causeDeath_filtered, y = resid, color = newStudy)) +
  geom_boxplot() +
  ylab('Residualized CTP Estimate') + 
  xlab('Cause of Death') + 
  theme(
    axis.text.x = element_text(size = 13, angle = 90),
    axis.text.y = element_text(size = 13),
    axis.title.x = element_text(vjust = -2, size = 16),
    axis.title.y = element_text(size = 16),
    strip.text.x = element_text(size = 16)) +
  facet_grid(cols = vars(cell_type), rows = vars(newStudy)) 
  #scale_y_continuous(limits = c(-2, 2)) 
resid_vs_ageonset



```


Let's make some more figures. Here's what we need to illustrate:
-Having SCZ when dying younger = less interneurons
-Dying by suicide = less interneurons 
-Dying by suicide and dying young w/ SCZ are almost completely confounded 
-Having SCZ and dying younger = more brain pathology
-Having SCZ and dying older = paradoxically less brain pathology 

Plot 1:
Re-do figure 2 from the main figures, but with only GVEX and LIBD 
-Model using controls and SCZ + Suicide deaths
-Model using controls and SCZ - Suicide deaths

```{r}
# Create function to map age to age range
age_to_range <- function(age) {
  if (age >= 90) {
    "90+"
  } else {
    lower_bound <- floor(age / 10) * 10
    upper_bound <- lower_bound + 9
    paste0(lower_bound, "-", upper_bound)
  }
}
#Create aggregated causeDeath column
all_estimations_long$age_range = sapply(all_estimations_long$ageDeath, age_to_range)
all_estimations_long = all_estimations_long %>%
    mutate(causeDeath_filtered = ifelse(grepl("SUIC", toupper(causeDeath)), "suicide",
                                        ifelse(grepl("DROWNING|STRANGULATION|INFECTION|ASPHYX|FALL|ACCIDENT|TRAUMA|MVA|OD|HOMIC|HEMORR", toupper(causeDeath)), "accident",
                                        ifelse(grepl("CARD|PULM", toupper(causeDeath)), "cardio",
                                        ifelse(grepl("NATURAL|CANCER|CIRROSIS|COPD|PNEU|SLEEP|PANCREAT|ASTHM", toupper(causeDeath)), "natural",
                                               "other")))))
```

```{r}
#Filter to include all controls, and only SCZ cases that did NOT die by suicide -> Cut off at ageDeath < 90 to have enough data for grouping
mega_lms_by_age_no_suicide = all_estimations_long %>% 
  filter(dataset %in% c("GVEX", "LIBD"), ageDeath < 50) %>% 
  filter(primaryDiagnosis == 'control' | (primaryDiagnosis == 'Schizophrenia' & causeDeath_filtered == "other")) %>%
  group_by(age_range, cell_type) %>%
  do(tidy(lm(scale(rel_prop) ~ scale(RIN) + scale(PMI) + reportedGender + primaryDiagnosis + newStudy, data = .))) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(`p.value`, method = 'BH')) %>%
  mutate(class = case_when(
    str_detect(cell_type, "Inh") ~ "Inhibitory",
    str_detect(cell_type, "Exc") ~ "Excitatory",
    TRUE ~ "Non-Neuronal")) %>%
  mutate(term = recode(term,
                       `(Intercept)` = "Intercept",
                       `reportedGendermale` = "gender:Male",
                       `primaryDiagnosisSchizophrenia` = "SCZ",
                       `scale(ageDeath)` = "ageDeath",
                       `scale(PMI)` = "PMI", 
                       `newStudyNIMH_HBCC` = "NIMH_HBCC",
                       `newStudyPitt` = "Pitt",
                       `newStudyMSSM` = "MSSM",
                       `newStudyPenn` = "Penn",
                       `scale(RIN)` = "RIN"))


#Plot it
mega_lms_by_age_no_suicide %>% 
  filter(term %in% 'SCZ' & cell_type %in% c("Inh_SST", "Inh_PVALB", "Inh_VIP")) %>% 
  mutate(cell_type = fct_reorder(cell_type, estimate)) %>% 
  ggplot(aes(x = age_range, y = estimate)) + 
  geom_hline(yintercept = 0) + 
  geom_bar(stat = "identity", position = "dodge") + 
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error)) + 
  ylab('Beta Coefficient') + 
  xlab('\n \n Age Range') +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 13), 
    axis.text.y = element_text(size = 13),
    axis.title.x = element_text(vjust = 8, size = 16),
    axis.title.y = element_text(size = 16),
    strip.text.x = element_text(size = 16)) +
  facet_wrap(~cell_type, drop = T, scale = "free_x") +
  guides(fill = "none") +
  scale_y_continuous(limits = c(-1.0, 1.4))

#This looks weird -> likely because removing CMC data removes a LOT of older subjects

```

In ageDeath <50, SCZ, and interneurons:
-Barplot CTP for suicide deaths vs. accident vs. cardiovascular 
-Scatterplot for CTP vs ageDeath in only suicide deaths 
```{r}
CELL_TYPES = c("Inh_SST", "Inh_PVALB", "Inh_VIP")
DATASETS = c("GVEX", "LIBD")


residuals_data = matrix(ncol = (ncol(all_estimations_long)+1), nrow = 0) %>% data.frame()
colnames(residuals_data) = c(colnames(all_estimations_long), "resid")
for(DATASET in DATASETS) {
  for(CELL_TYPE in CELL_TYPES) {
    #Create res_model -> no ageDeath
    res_model_data = all_estimations_long %>% filter(cell_type == CELL_TYPE , dataset == DATASET, primaryDiagnosis %in% c("control", "Schizophrenia"))
    res_model_data$primaryDiagnosis = res_model_data$primaryDiagnosis %>% factor(levels = c("control", "Schizophrenia"))
    res_model = lm(scale(rel_prop) ~ primaryDiagnosis + scale(ageDeath) + scale(RIN) + scale(PMI) + reportedGender, data = res_model_data)
    #Calculate residuals and add column to res_model_data
    res_model_data = res_model_data %>% add_residuals(var = "resid", model = res_model)
    residuals_data = residuals_data %>% rbind(res_model_data)
  }
}

residuals_data %>% filter() %>%
  ggplot(aes(x = causeDeath_filtered, y = resid)) +
  geom_boxplot(outlier.shape = NA, show.legend = F, notch = F, aes(fill = causeDeath_filtered, alpha = 0.5)) +
  geom_beeswarm(size = 3, alpha = 0.3, aes(colour = causeDeath_filtered), show.legend = F) +
  facet_grid(cols = vars(cell_type)) +
  geom_signif(comparisons = combn(c("suicide", "accident", "natural", "cardio"), 2) %>% t() %>% split(1:4), map_signif_level = TRUE)



all_estimations_long %>%
mutate(causeDeath_filtered = ifelse(grepl("SUIC", toupper(causeDeath)), "suicide", "other")) %>%
 filter(primaryDiagnosis %in% c('Schizophrenia', 'control'), 
        cell_type %in% c("Inh_SST", "Inh_PVALB", "Inh_VIP"), 
        dataset %in% c("GVEX", "LIBD")) %>%
  ggplot(aes(x = ageDeath, y = rel_prop, color = causeDeath_filtered)) +
  geom_point(size = 0.5) +
  geom_smooth(method = lm, se = F) +
  scale_y_continuous(limits = c(-2.5, 2.5)) +
  facet_grid(cols = vars(primaryDiagnosis), rows = vars(dataset))
  
```

What have we learned?
-It's basically impossible to disentangle young ageDeath from death by suicide 
-AgeDeath and CauseDeath are related, 









