---
title: "psychencode_figure_3"
author: "Daniel Kiss"
date: "`r Sys.Date()`"
output: html_document
---
Packages etc.
```{r}
library(dplyr)
library(tidyverse)
library(matrixStats)
library(cowplot)
library(broom)
library(knitr)
library(ggpubr)
library(ggrepel)
library(ggbeeswarm)
library(patchwork)
library(ggsignif)
library(modelr)
theme_set(theme_classic2())
#Colour palette
cbPalette = c("#56B4E9", "#009E73","#E69F00", "#0072B2", "#D55E00", "#CC79A7","#000000","#F0E442")
```

Load Data
```{r}
mgp_estimations = read_csv("Desktop/Triplab/PsychAD /psychencode_mgp_estimations.csv")
#Cell type proportions for each data set merged with metadata - ready for modelling 
mgp_estimations = mgp_estimations %>% filter(ageDeath >= 15)
mgp_estimations$newStudy = mgp_estimations$newStudy %>% factor(levels = c("Pitt", "GVEX", "NIMH_HBCC", "LIBD_szControl", "MSSM", "Penn"))
mgp_estimations_long = mgp_estimations %>% pivot_longer(cols = Astrocyte:VLMC, names_to = 'cell_type', values_to = 'rel_prop') %>%
  mutate(cell_class = case_when(
    cell_type %in% c( "PVALB", "SST", "VIP", "LAMP5", "PAX6") ~ "Inhibitory",
    cell_type %in% c("IT", "L4.IT", "L5.ET", "L5.6.IT.Car3", "L5.6.NP", "L6.CT", "L6b") ~ "Excitatory",
    TRUE ~ "Non-Neuronal"
  ))
mgp_estimations_long$cell_type = mgp_estimations_long$cell_type %>% factor(levels = c("PVALB", "SST", "VIP", "LAMP5", "PAX6", "IT", "L4.IT", "L5.ET", "L5.6.IT.Car3", "L5.6.NP", "L6.CT", "L6b", "Astrocyte", "Endothelial", "Microglia", "Oligodendrocyte", "OPC", "Pericyte", "VLMC"))
mgp_estimations_long$cell_class = mgp_estimations_long$cell_class %>% factor(levels = c("Inhibitory", "Excitatory", "Non-Neuronal"))
```

Residualize MGPs based on linear model
```{r}
mgp_estimations_long = mgp_estimations_long %>%
  group_by(cell_type) %>%
  mutate(resid = resid(lm(scale(rel_prop) ~ scale(RIN) + scale(ageDeath) + scale(PMI) + 
                reportedGender))) %>%
  ungroup()
```

Model rCTPs based on diagnosis and covariates seperately in each cohort
```{r}
combined_lms = mgp_estimations_long %>%
  group_by(newStudy, cell_type) %>%
  do(tidy(lm(scale(rel_prop) ~ scale(RIN) + scale(ageDeath) + scale(PMI) +
               reportedGender + primaryDiagnosis, data = .))) %>%
  ungroup() %>%
  mutate(padj = p.adjust(p.value, method = 'fdr'),
         term = recode(term,
                       `(Intercept)` = "Intercept",
                       `reportedGendermale` = "gender:Male",
                       `primaryDiagnosisSchizophrenia` = "SCZ",
                       `scale(ageDeath)` = "Age",
                       `scale(PMI)` = "PMI",
                       `scale(RIN)` = "RIN")) %>%
  merge(unique(mgp_estimations_long[c('dataset', 'newStudy')]), by.x = "newStudy") %>%
  merge(unique(mgp_estimations_long[c('cell_class', 'cell_type')]), by.x = "cell_type") %>%
  mutate(newStudy = factor(newStudy, levels = c("Pitt", "GVEX", "NIMH_HBCC", "LIBD_szControl", "MSSM", "Penn")))
```

Interaction Models: As outlined above, there appears to be an interaction between age x diagnosis with respect to SST cell proportion. In other words, SST cell proportion is generally lower as ageDeath increases, but declines faster with age in controls. To analyze this more clearly, we create new models (interaction_lms) to visualize this effect. These models include new_study as a covariate, and age x diagnosis as an interactor.
```{r}
interaction_lms = mgp_estimations_long %>%
  group_by(newStudy, cell_type) %>%
  do(tidy(lm(scale(rel_prop) ~ scale(RIN) + scale(ageDeath) + scale(PMI) +
               reportedGender + primaryDiagnosis + scale(ageDeath) * primaryDiagnosis, data = .))) %>%
  ungroup() %>%
  mutate(padj = p.adjust(`p.value`, method = 'fdr'),
         term = recode(term,
                       `(Intercept)` = "Intercept",
                       `reportedGendermale` = "gender:Male",
                       `primaryDiagnosisSchizophrenia` = "SCZ",
                       `scale(ageDeath):primaryDiagnosisSchizophrenia` = "Age x Diagnosis",
                       `scale(ageDeath)` = "Age",
                       `scale(PMI)` = "PMI",
                       `scale(RIN)` = "RIN")) %>%
  merge(unique(mgp_estimations_long[c('dataset', 'newStudy')]), by.x = "newStudy") %>%
  merge(unique(mgp_estimations_long[c('cell_class', 'cell_type')]), by.x = "cell_type") %>%
  mutate(newStudy = factor(newStudy, levels = c("Pitt", "GVEX", "NIMH_HBCC", "LIBD_szControl", "MSSM", "Penn")))

```

Interaction mega-analysis w/ all data 
```{r}
mega_lms = mgp_estimations_long %>% 
  group_by(cell_type) %>%
  do(tidy(lm(scale(rel_prop) ~ scale(RIN) + scale(ageDeath) + scale(PMI) +
             reportedGender + primaryDiagnosis + newStudy + scale(ageDeath)*primaryDiagnosis, data = .))) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(`p.value`, method = 'fdr'),
         term = recode(term,
                       `(Intercept)` = "Intercept",
                       `reportedGendermale` = "gender:Male",
                       `primaryDiagnosisSchizophrenia` = "SCZ",

                       `scale(ageDeath)` = "Age",
                       `scale(ageDeath):primaryDiagnosisSchizophrenia` = "Age x Diagnosis",
                       `scale(PMI)` = "PMI", 
                       `newStudyNIMH_HBCC` = "NIMH_HBCC",
                       `newStudyLIBD_szControl` = "LIBD_szControl",
                       `newStudyPitt` = "Pitt",
                       `newStudyMSSM` = "MSSM",
                       `newStudyPenn` = "Penn",
                       `scale(RIN)` = "RIN")) %>%
  merge(unique(mgp_estimations_long[c('cell_class', 'cell_type')]), by.x = "cell_type") 
```

Functions to  generate age range bins 
```{r}
# Create function to map age to age range
age_to_range = function(age) {
    if(age >= 15 & age < 30) {
    return("15-29")
  } else {
    lower_bound = floor((age - 30) / 10) * 10 + 30
    upper_bound = lower_bound + 9
    return(paste0(lower_bound, "-", upper_bound))
  }
}
# Create function to bin ages to >=70 and <70
classify_ages = function(ages) {
  classifications = ifelse(ages <= 70, "Under 70", "Over 70")
  return(classifications)
}
```


Create age groups from age 20-90 in increments of 10 years. Plot meganalysis beta coeffs against age group.
```{r}
mgp_estimations_long$age_range = sapply(mgp_estimations_long$ageDeath, age_to_range)

mega_lms_by_age = mgp_estimations_long %>% 
  group_by(age_range, cell_type) %>%
  do(tidy(lm(scale(rel_prop) ~ scale(RIN) + scale(PMI) +
             reportedGender + primaryDiagnosis + newStudy, data = .))) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(`p.value`, method = 'fdr'),
         term = recode(term,
                       `(Intercept)` = "Intercept",
                       `reportedGendermale` = "gender:Male",
                       `primaryDiagnosisSchizophrenia` = "SCZ",
                       `scale(ageDeath)` = "Age",
                       `scale(PMI)` = "PMI", 
                       `newStudyNIMH_HBCC` = "NIMH_HBCC",
                       `newStudyPitt` = "Pitt",
                       `newStudyMSSM` = "MSSM",
                       `newStudyPenn` = "Penn",
                       `scale(RIN)` = "RIN")) %>%
  merge(unique(mgp_estimations_long[c('cell_class', 'cell_type')]), by.x = "cell_type") 
```

Bulk RNAseq Meganalysis based on age +/- 70
```{r}
mgp_estimations_long$age_class <- sapply(mgp_estimations_long$ageDeath, classify_ages) %>% factor(levels = c("Under 70", "Over 70"))
mega_lms_by_age_class = mgp_estimations_long %>% 
  group_by(age_class, cell_type) %>%
  do(tidy(lm(scale(rel_prop) ~ scale(RIN) + scale(PMI) + scale(ageDeath) + 
             reportedGender + newStudy + primaryDiagnosis, data = .))) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(`p.value`, method = 'fdr'),
         term = recode(term,
                       `(Intercept)` = "Intercept",
                       `reportedGendermale` = "gender:Male",
                       `primaryDiagnosisSchizophrenia` = "SCZ",
                       `scale(ageDeath)` = "Age",
                       `scale(PMI)` = "PMI", 
                       `newStudyNIMH_HBCC` = "NIMH_HBCC",
                       `newStudyPitt` = "Pitt",
                       `newStudyMSSM` = "MSSM",
                       `newStudyPenn` = "Penn",
                       `scale(RIN)` = "RIN")) %>%
  merge(unique(mgp_estimations_long[c('cell_class', 'cell_type')]), by.x = "cell_type") 

#same but stratify by cohort
lms_by_age_class = mgp_estimations_long %>% 
  group_by(age_class, cell_type, newStudy) %>%
  do(tidy(lm(scale(rel_prop) ~ scale(RIN) + scale(PMI) + scale(ageDeath) +
             reportedGender + primaryDiagnosis, data = .))) %>%
  ungroup() %>% 
  mutate(padj = p.adjust(`p.value`, method = 'fdr'),
         term = recode(term,
                       `(Intercept)` = "Intercept",
                       `reportedGendermale` = "gender:Male",
                       `primaryDiagnosisSchizophrenia` = "SCZ",
                       `scale(ageDeath)` = "Age",
                       `scale(PMI)` = "PMI", 
                       `newStudyNIMH_HBCC` = "NIMH_HBCC",
                       `newStudyPitt` = "Pitt",
                       `newStudyMSSM` = "MSSM",
                       `newStudyPenn` = "Penn",
                       `scale(RIN)` = "RIN")) %>%
  merge(unique(mgp_estimations_long[c('cell_class', 'cell_type')]), by.x = "cell_type") 
```


```{r}
# Filter combined_lms for the desired cell types and term
combined_filtered <- lms_by_age_class %>%
  filter(cell_type %in% c("PVALB", "SST", "VIP"), term == "SCZ")

# Filter mega_lms for the same conditions and add newStudy column as "All"
mega_filtered <- mega_lms_by_age_class %>%
  filter(cell_type %in% c("PVALB", "SST", "VIP"), term == "SCZ") %>%
  mutate(newStudy = "Mega")

# Combine the filtered data frames into mega_results
mega_results <- bind_rows(combined_filtered, mega_filtered) %>%
  rename(Cohort = newStudy)

mega_results <- mega_results %>%
  mutate(CI_lower = estimate - 1.96 * std.error,
         CI_upper = estimate + 1.96 * std.error)


```


```{r}
# Ensure the original dataset has the age_class column
mgp_estimations <- mgp_estimations %>%
  mutate(age_class = sapply(ageDeath, classify_ages))

# Compute sample sizes by cohort (newStudy) and age_class
sample_sizes <- mgp_estimations %>%
  group_by(newStudy, age_class) %>%
  summarise(Sample_Size = n()) %>%
  ungroup()

# Compute sample sizes for the mega-analysis ("All") for each age_class
sample_sizes_all <- mgp_estimations %>%
  group_by(age_class) %>%
  summarise(Sample_Size = n()) %>%
  mutate(newStudy = "Mega") %>%
  ungroup()

# Combine the sample sizes
sample_sizes_combined <- bind_rows(sample_sizes, sample_sizes_all)

# Join the sample sizes to mega_results (matching on age_class and Cohort)
mega_results <- mega_results %>%
  left_join(sample_sizes_combined, by = c("Cohort" = "newStudy", "age_class"))

```


```{r, fig.width=12, fig.height=4}
# Step 1: Define custom shape mapping
mega_results <- mega_results %>%
  mutate(
    Shape = ifelse(Cohort == "Mega", "Pooled", "Individual")
  )

# Step 2: Factor Cohort to control order
mega_results$Cohort <- factor(mega_results$Cohort,
                             levels = c("Mega", "Penn", "MSSM", "LIBD_szControl", "NIMH_HBCC", "GVEX", "Pitt"))

mega_results$AgeGroup <- factor(mega_results$age_class, levels = c("Under 70", "Over 70"))

mega_results_sig <- mega_results %>% 
  mutate(
    ci_excludes_zero = CI_lower > 0 | CI_upper < 0,
    # aterisk string based on FDR-adjusted P
    signif_lab = case_when(
      ci_excludes_zero & padj <= 0.01 ~ "***",
      ci_excludes_zero & padj <= 0.05 ~ "**",
      ci_excludes_zero & padj <= 0.10 ~ "*",
      TRUE                            ~ ""
    ),
    signif_x   = ifelse(estimate > 0, CI_upper, CI_lower),
    hjust_val  = ifelse(estimate > 0, -0.25, 1.25)
  )

# Step 3: Plot
fp <- ggplot(mega_results_sig, aes(x = estimate, y = Cohort)) +
  geom_point(aes(size = Sample_Size, color = Cohort, shape = Shape)) +
  geom_errorbarh(aes(xmin = CI_lower, xmax = CI_upper, color = Cohort),
                 height = 0.2) +
  geom_text(
    aes(x = signif_x, label = signif_lab, hjust = hjust_val),
    vjust = 0.5,                                 # centred vertically on the row
    size  = 5, fontface = "bold", na.rm = TRUE
  ) +
  facet_grid(AgeGroup ~ cell_type, switch = "y") +
  scale_shape_manual(values = c("Pooled" = 18, "Individual" = 16)) +
  scale_size_continuous(range = c(3, 8)) +
  scale_color_manual(values = c(
    "Pitt"          = "#56B4E9",
    "GVEX"          = "#009E73",
    "NIMH_HBCC"     = "#E69F00",
    "LIBD_szControl"= "#0072B2",
    "MSSM"          = "#D55E00",
    "Penn"          = "#CC79A7",
    "Mega"           = "#808080"
  )) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x = "Effect Size (Estimate)",
    y = "Cohort",
    size = "Sample Size"
  ) +
  theme_classic() +
  theme(
    legend.position = "right",
    axis.text.x     = element_text(size = 15),
    axis.text.y     = element_blank(),
    axis.title.x    = element_text(vjust = -2, size = 15),
    axis.title.y    = element_text(size = 15),
    legend.title    = element_blank(),
    legend.text     = element_text(size = 12),
    strip.text.x    = element_text(size = 15),
    strip.text.y    = element_text(size = 15),
    plot.margin     = margin(t = 5, r = 15, b = 15, l = 5)   # avoids clipping
  ) +
  guides(shape = "none")

fp

fp

```

```{r, fig.height=16, fig.width=12}
# Plot interaction beta values from meganalysis
significant_mega = subset(mega_lms, padj <= 0.1 & term == "Age x Diagnosis" & cell_type %in% c("Astrocyte", "IT", "L5.6.NP", "LAMP5", "Microglia", "Oligodendrocyte", "OPC", "Pericyte", "PVALB", "SST", "VIP" ))
significant_mega$asterisks = ifelse(significant_mega$padj <= 0.01, "***",
                                   ifelse(significant_mega$padj <= 0.05, "**", "*"))
significant_mega$vjust = ifelse(significant_mega$estimate >= 0, -3, 3)

# Figure 2a: Results from interaction meganalysis stratified by cell class - plot interaction beta coeff
mega_lms$cell_type = mega_lms$cell_type %>% factor(levels = c("PVALB", "SST", "VIP", "LAMP5", "PAX6", "IT", "L4.IT", "L5.ET", "L5.6.IT.Car3", "L5.6.NP", "L6.CT", "L6b", "Astrocyte", "Endothelial", "Microglia", "Oligodendrocyte", "OPC", "Pericyte", "VLMC"))
figure_2a = mega_lms %>%  filter(cell_type %in% c("Astrocyte", "IT", "L5.6.NP", "LAMP5", "Microglia", "Oligodendrocyte", "OPC","Pericyte", "PVALB", "SST", "VIP" )) %>%
  filter(term %in% 'Age x Diagnosis') %>% 
  ggplot(aes(x = cell_type, y = estimate)) + 
  geom_hline(yintercept = 0) + 
  geom_bar(stat = "identity", position = "dodge") + 
  facet_grid(~cell_class, drop = T, scale = "free_x", space = "free") +
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error)) + 
  ylab('Interaction Coefficient') + 
  xlab('\n \n Cell Type') + 
  geom_text(data = significant_mega,
            aes(cell_type, estimate, label = asterisks), 
            vjust = significant_mega$vjust) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 12), 
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(vjust = 8, size = 15),
    axis.title.y = element_text(size = 15),
    strip.text.x = element_text(size = 15)) +
  guides(fill = FALSE) +
  scale_y_continuous(limits = c(-0.15, 0.32))

# Figure 2b: Plot beta values from meganalysis in 10-year age bins for SST, PVALB, VIP
significant_mega_age = subset(mega_lms_by_age, padj <= 0.1 & term == "SCZ" & cell_type %in% c("SST", "PVALB", "VIP"))
significant_mega_age$asterisks = ifelse(significant_mega_age$padj <= 0.01, "***",
                                         ifelse(significant_mega_age$padj <= 0.05, "**", "*"))
significant_mega_age$vjust = ifelse(significant_mega_age$estimate >= 0, -2.5, 2.5)

figure_2b = mega_lms_by_age %>% 
  filter(term %in% 'SCZ' & cell_type %in% c("SST", "PVALB", "VIP")) %>% 
  mutate(cell_type = fct_reorder(cell_type, estimate)) %>% 
  ggplot(aes(x = age_range, y = estimate)) + 
  geom_hline(yintercept = 0) + 
  geom_bar(stat = "identity", position = "dodge") + 
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error)) + 
  ylab('Beta Coefficient') + 
  xlab('\n \n Age Range') + 
   geom_text(
    data   = significant_mega_age,
    aes(age_range, estimate, label = asterisks, vjust = vjust),
    size   = 5,       
    fontface = "bold"  
  ) +
   theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 13), 
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(vjust = 8, size = 15),
    axis.title.y = element_text(size = 15),
    strip.text.x = element_text(size = 15)) +
  facet_wrap(~cell_type, drop = T, scale = "free_x") +
  guides(fill = FALSE) +
  scale_y_continuous(limits = c(-1.1, 1.7))


# Figure 2c: Visualizing interaction between age and diagnosis by plotting rel_prop vs. age in each diagnosis 
figure_2c = mgp_estimations_long %>% filter(ageDeath < 90) %>%
  filter(cell_type %in% c("SST", "PVALB", "VIP")) %>%
  ggplot(aes(x = ageDeath, y = rel_prop, color = primaryDiagnosis)) +
  geom_point(alpha = 0.5, size = 0.5) + 
  geom_smooth(se = F, method = 'lm', fullrange = T) +
  ylab('Relative Cell Prop.') + 
  xlab('Age') + 
  theme(
    axis.text.x = element_text(size = 15),
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(vjust = -2, size = 15),
    axis.title.y = element_text(size = 15),
    legend.position = c(0.9, 0.9),
    legend.title = element_blank(),
    legend.text = element_text(size = 12),  
    strip.text.x = element_text(size = 15)) +
  facet_grid(~cell_type, drop = T, scale = "free_x", space = "free") +
  guides(fill = guide_legend(nrow = 1)) +
  scale_y_continuous(limits = c(-3, 3)) + 
  scale_color_manual(values = c("dodgerblue2", "firebrick2"), labels=c('CON', 'SCZ'))

#Figure 2d: Mega-analysis results when binned as +/- 70 years
significant_mega_age_class = subset(mega_lms_by_age_class, padj <= 0.1 & term == "SCZ" & cell_type %in% c("SST", "PVALB", "VIP"))
significant_mega_age_class$asterisks = ifelse(significant_mega_age_class$padj <= 0.01, "***", ifelse(significant_mega_age_class$padj <= 0.05, "**", "*"))
significant_mega_age_class$vjust = ifelse(significant_mega_age_class$estimate >= 0, -3, 3)

figure_2d = mega_lms_by_age_class %>% 
  filter(term %in% 'SCZ' & cell_type %in% c("SST", "PVALB", "VIP")) %>% 
  mutate(cell_type = fct_reorder(cell_type, estimate)) %>% 
  ggplot(aes(x = age_class, y = estimate, fill = age_class)) + 
  geom_hline(yintercept = 0) + 
  geom_bar(stat = "identity", position = "dodge") + 
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error)) + 
  ylab('Beta Coefficient') + 
  xlab('\n \n Age Range') + 
  scale_fill_manual(values = c("mediumpurple3", "olivedrab3"), labels=c('Above 70', 'Below 70')) +
  geom_text(data = significant_mega_age_class,
            aes(age_class, estimate, label = asterisks), 
            vjust = significant_mega_age_class$vjust) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 13), 
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(vjust = 8, size = 15),
    axis.title.y = element_text(size = 15),
    strip.text.x = element_text(size = 15)) +
  facet_wrap(~cell_type, drop = T, scale = "free_x") +
  guides(fill = FALSE) +
  scale_y_continuous(limits = c(-0.55, 0.55))

#   All plots get a small margin to stop label clipping
margin_fix <- theme(plot.margin = margin(b = 10))

figure_2c <- figure_2c + margin_fix
figure_2b <- figure_2b 
fp        <- fp        

figure_2 <- (
    figure_2c /         
    figure_2b /         
    fp                  
) +
  plot_layout(heights = c(1, 1, 1.8)) +        
  plot_annotation(tag_levels = 'A') &        
  theme(
    plot.tag      = element_text(size = 25),
  )

figure_2   
```





