---
title: "psychencode_DE_file_generation"
author: "Daniel Kiss"
date: "8/16/2023"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(Seurat)
  library(edgeR)
  library(ggrepel)
  library(cowplot)
  library(limma)
  library(ggpubr)
  library(gprofiler2)
  library(EnhancedVolcano)
  library(patchwork)
})
theme_set(theme_classic2())
#Colour palette
cbPalette <- c("#56B4E9", "#009E73","#E69F00", "#0072B2", "#D55E00", "#CC79A7","#000000","#F0E442")
```


Load and format data/variables
```{r}
pseudobulk = read.csv("~/cell_prop_psychiatry/data/psychencode_label_transferred_pseudobulk_raw.csv")[,-1]
pseudobulk_metadata = read.csv("~/cell_prop_psychiatry/data/psychencode_label_transferred_pseudobulk_metadata.csv") %>%
  group_by(ID) %>%
  mutate(total_cells = sum(num_cells)) %>%
  ungroup() 
```


Create a list of relevant data for each cell type
```{r}
#Define variables
NUM_CELLS = 1000 #Total cells per donor in all cell types
AGE_THRESH = 70 
CELL_TYPES = c("SST", "PVALB", "VIP")
DATASET = c("McLean", "MtSinai")

DE_data = list()
#Loop through ALL cell types and perform above analyses
for(CELL_TYPE in CELL_TYPES) {
  #Define subjects based on cell type, cohort, and min cell number
  under_70_subjects = pseudobulk_metadata %>% filter(Celltype == CELL_TYPE, total_cells > NUM_CELLS, Age <= 70, Cohort %in% DATASET) %>% 
  pull(ID)
  over_70_subjects = pseudobulk_metadata %>% filter(Celltype == CELL_TYPE, total_cells > NUM_CELLS, Age > 70, Cohort %in% DATASET) %>% 
  pull(ID)
  #Define filtering variables 
  UNDER_70_SCZ = pseudobulk_metadata %>% filter(ID %in% under_70_subjects, Celltype == CELL_TYPE, Phenotype == "SZ") %>% pull(ID)
  UNDER_70_CON = pseudobulk_metadata %>% filter(ID %in% under_70_subjects, Celltype == CELL_TYPE, Phenotype == "CON") %>% pull(ID)
  OVER_70_SCZ = pseudobulk_metadata %>% filter(ID %in% over_70_subjects, Celltype == CELL_TYPE, Phenotype == "SZ") %>% pull(ID)
  OVER_70_CON = pseudobulk_metadata %>% filter(ID %in% over_70_subjects, Celltype == CELL_TYPE, Phenotype == "CON") %>% pull(ID)
  #Skip cell type if no subjects make it past filtering
  if(length(UNDER_70_SCZ) < 5 || length(UNDER_70_CON) < 5 || length(OVER_70_SCZ) < 5 || length(OVER_70_CON) < 5) {
    next  # Skip to the next iteration of the loop
  }
  #Define model variables for under 70
  disease = factor( pseudobulk_metadata %>% filter(ID %in% under_70_subjects, Celltype == CELL_TYPE) %>% 
    pull(Phenotype), levels = c('CON', 'SZ'))
  sex =  pseudobulk_metadata %>% filter(ID %in% under_70_subjects, Celltype == CELL_TYPE) %>% 
    pull(Gender)
  pmi =  pseudobulk_metadata %>% 
  filter(ID %in% under_70_subjects, Celltype == CELL_TYPE) %>% 
  pull(PMI)
  age =  pseudobulk_metadata %>% 
  filter(ID %in% under_70_subjects, Celltype == CELL_TYPE) %>% 
  pull(Age)
  log_10_cells_per_donor =  pseudobulk_metadata %>% 
  filter(ID %in% under_70_subjects, Celltype == CELL_TYPE) %>% 
  pull(num_cells) %>% log10()
  cohort = if (length(unique(DATASET)) > 1) pseudobulk_metadata %>% 
  filter(ID %in% under_70_subjects, Celltype == CELL_TYPE) %>% 
  pull(Cohort) else NULL
  #Create model matrix 
  design_under_70 = model.matrix(~ age + pmi + sex + log_10_cells_per_donor + disease) 
  
    #Define model variables for over 70
  disease = factor( pseudobulk_metadata %>% filter(ID %in% over_70_subjects, Celltype == CELL_TYPE) %>% 
    pull(Phenotype), levels = c('CON', 'SZ'))
  sex =  pseudobulk_metadata %>% filter(ID %in% over_70_subjects, Celltype == CELL_TYPE) %>% 
    pull(Gender)
  pmi =  pseudobulk_metadata %>% 
  filter(ID %in% over_70_subjects, Celltype == CELL_TYPE) %>% 
  pull(PMI)
  age =  pseudobulk_metadata %>% 
  filter(ID %in% over_70_subjects, Celltype == CELL_TYPE) %>% 
  pull(Age)
  log_10_cells_per_donor =  pseudobulk_metadata %>% 
  filter(ID %in% over_70_subjects, Celltype == CELL_TYPE) %>% 
  pull(num_cells) %>% log10()
  cohort = if (length(unique(DATASET)) > 1) pseudobulk_metadata %>% 
  filter(ID %in% over_70_subjects, Celltype == CELL_TYPE) %>% 
  pull(Cohort) else NULL
  #Create model matrix 
  design_over_70 = model.matrix(~age + pmi + sex + log_10_cells_per_donor +  disease) 
  
  #Create count matrices
  under_70 = pseudobulk_metadata %>% filter(Age <= 70, ID %in% under_70_subjects, Celltype == CELL_TYPE) %>% pull(ID) 
  counts_under_70 = pseudobulk %>% filter(ID %in% under_70, Celltype == CELL_TYPE)
  counts_under_70 = counts_under_70[, 3:ncol(counts_under_70)] %>% t() %>% as.data.frame()
  colnames(counts_under_70) = under_70
  over_70 = pseudobulk_metadata %>% filter(Age > 70, ID %in% over_70_subjects, Celltype == CELL_TYPE) %>% pull(ID) 
  counts_over_70 = pseudobulk %>% filter(ID %in% over_70, Celltype == CELL_TYPE)
  counts_over_70 = counts_over_70[, 3:ncol(counts_over_70)] %>% t() %>% as.data.frame()
  colnames(counts_over_70) = over_70
  #Create DGE objects
  dge_under = DGEList(counts_under_70, genes = row.names(counts_under_70))
  min_samples <- pseudobulk_metadata %>% filter(ID %in% under_70_subjects, Celltype == CELL_TYPE) %>% nrow * 0.8
  dge_under = dge_under[rowSums(dge_under$counts >= 1) >= min_samples, ]
  dge_under = calcNormFactors(dge_under, method = "TMM")
  dge_over = DGEList(counts_over_70, genes = row.names(counts_over_70))
  min_samples <- pseudobulk_metadata %>% filter(ID %in% over_70_subjects, Celltype == CELL_TYPE) %>% nrow * 0.8
  dge_over = dge_over[rowSums(dge_over$counts >= 1) >= min_samples, ]
  dge_over = calcNormFactors(dge_over, method = "TMM")
  
  #Create lists of data for each group
  under_70_data = list(design_under_70, dge_under)
  names(under_70_data) = c("design", "dge")
  over_70_data = list(design_over_70, dge_over)
  names(over_70_data) = c("design", "dge")
  joined_data = list(under_70_data, over_70_data)
  names(joined_data) = c("under_70", "over_70")
  
  DE_data[[CELL_TYPE]] = joined_data
}
#We now have a nested list with ALL data required for DE analysis for all cell types with >1 subject per age group
```

Simple volcano plot function 
```{r}
simple_volcano <- function(x) {
  x %>%
    ggplot(aes(x = logFC, y = -log10(P.Value), label = genes)) + 
    geom_point() +
    geom_text_repel(data = subset(x, adj.P.Val < 0.1), 
                    aes(label = genes), 
                    vjust = 1.5) + 
    geom_point(data = subset(x, adj.P.Val < 0.1), 
               aes(color = 'red')) + 
    theme_cowplot() +
    theme(legend.position = "none")
}
```


Now that we have all data pre-processed and organized for DE analysis, let's loop through DE_data and output results to DE_results (This will take a few minutes to run)
```{r}
DE_results = list()
for(CELL_TYPE in names(DE_data)) {
  #Load data from DE_data
  dge_under = DE_data[[CELL_TYPE]]$under_70$dge 
  dge_over = DE_data[[CELL_TYPE]]$over_70$dge 
  design_under = DE_data[[CELL_TYPE]]$under_70$design
  design_over = DE_data[[CELL_TYPE]]$over_70$design
  #Voom transform
  vm_under = voom(dge_under, design_under, plot = F) 
  fit_under = lmFit(vm_under, design_under) %>% eBayes()
  DE_under = topTable(fit_under, coef = "diseaseSZ",  n = Inf, sort = "none", adjust.method = "BH", confint = T) 
  vm_over = voom(dge_over, design_over, plot = FALSE) 
  fit_over = lmFit(vm_over, design_over) %>% eBayes()
  DE_over = topTable(fit_over, coef = "diseaseSZ", n = Inf, sort = "none", adjust.method = "BH", confint = T)
  #Plots etc.
  volcano_under = DE_under %>% simple_volcano()
  volcano_over = DE_over %>% simple_volcano()
  tplot = merge(DE_under, DE_over, by = 0) %>% 
    ggplot(aes(x = t.x, y = t.y)) + 
    geom_smooth(method = loess) +
    geom_point(size = 1) +
    ylab("Over 70") +
    xlab("Under 70") +
    ggtitle(paste0(CELL_TYPE, " t-Statistics"))
  
  #Append to list
  joined_results = list(DE_under, DE_over)
  names(joined_results) = c("DE_under", "DE_over")
  joined_volcano = list(volcano_under, volcano_over)
  names(joined_volcano) = c("Under 70", "Over 70")
  joined_all = list(joined_results, joined_volcano, tplot)
  names(joined_all) = c("DE_tables", "Volcano_Plots", "t_Plot")
  DE_results[[CELL_TYPE]] = joined_all
}
```

We will run the code above 2x, once for McLean and once for Sinai, then save each result for later use 
```{r}
dataset_suffix = if(length(DATASET) > 1) "ALL" else DATASET
saveRDS(DE_results, file = paste0("/external/rprshnas01/kcni/dkiss/cell_prop_psychiatry/data/DE_results_", dataset_suffix))
```
